### 2017-12-10 readme.rmd
first draft

---
output: 
  github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE,
	echo=FALSE
	)
```

# How are ctenophores and sponges related to other animals?

#### Casey W. Dunn, Department of Ecology and Evolutionary Biology, Yale University, http://dunnlab.org

There has been great interest in resolving the deepest evolutionary relationships in the animal tree of life. This comes down to resolving the relationships between five clades - Cnidaria, Bilateria, Placozoa, Porifera (sponges), and Ctenophora (comb jellies). The deepest split in the animal tree has been of particular interest, with most recent studies finding either Ctenophora or Porifera to be the sister group to all other animals. These are referred to as the Ctenophora-sister and Porifera-sister hypotheses (Figure 1).

![overview](figures/figure_overview.png)

*Figure 1- Overview. The two alternative hypotheses for deep animal relationships considered here. In the Ctenophora-sister hypothesis ctenophores are the sister group to all other animals, i.e. there is an internal branch in the animal phylogeny that separates ctenophores from all other animals (shown in red). In the Porifera-sister hypothesis sponges are the sister group to all other animals, i.e. there is an internal branch in the animal phylogeny that separates sponges from all other animals (shown in green). Other relationships that are not part of these hypotheses are shown as an unresolved polytomy. Figure adapted from [Dunn et al. 2014](http://www.annualreviews.org/doi/10.1146/annurev-ecolsys-120213-091627). Colors are based on the graphical abstract of Feuda et al. 2017.*

Over a series of studies from the past 10 years, it has come to be well known that the phylogenetic positions of ctenophores and sponges are highly sensitive to model of molecular evolution, species sampling, and gene sampling (reviewed in [Dunn *et al.* 2014](http://www.annualreviews.org/doi/10.1146/annurev-ecolsys-120213-091627)). In the course of exploring this sensitivity, some studies have concluded that there is strong support for Porifera-sister, some have concluded that there is strong support for Ctenophora-sister, and some have concluded there is not yet enough evidence to definitively reject either hypothesis. My own take on the question falls into this last category.

In a recent paper, [Feuda *et al.* 2017](https://doi.org/10.1016/j.cub.2017.11.008) claim to present strong significant support for Porifera-sister and to have resolved the question at last. Feuda *et al.* present a well-blocked set of phylogenetic analyses that examine support for Ctenophora-sister and Porifera-sister under a variety of conditions. They consider several factors:

- Model of molecular evolution. The primary analyses consider the GTR+G and CAT-GTR+G models. Both models accommodate different rates of change between states (General Time Reversible, GTR) and among-site rate variation (Gamma Distribution of rates, G). In addition, CAT-GTR+G accommodates different state frequencies across sites (Categorical model, CAT).

- Dataset. They examine two matrices from previous publications, [Whelan *et al.* 2015](https://www.doi.org/10.1073/pnas.1503453112) and [Chang *et al.* 2015](https://www.doi.org/10.1073/pnas.1511468112).

- Outgroup sampling. For the Whelan matrix, they consider three sets of species sampling with progressively reduced outgroups - Opisthokonta, Holozoa, and Choanimalia. They do not vary outgroup sampling for the Chang matrix, which includes species from Holozoa. Removing outgroups is a form of data reduction.

- Character recoding. They recoded amino acids with similar chemical properties as the same character using three different schemes (KGB-6, Dayhoff-6, SR-6). This recoding is a form of data reduction - rather than remove species or sites, it removes dynamic range within each cell of the matrix. The motivation of this recoding is to minimize the impacts of compositional bias, since different lineages and genes may tend to use different amino acids with similar properties.

This blocking of 2 models for 4 matrices for 4 types of character coding results in 32 analyses that form the central results of their study.

They interpret these results as providing strong significant support for Porifera-sister.

Here are the highlights that accompany their paper:

> Highlights
>
> This is the first comparison of model adequacy in non- bilaterian phylogenomics
>
> Animal relationships depend on the ability of the models to describe the data
>
> Sponges are the sister group of the remaining animals
>
> Trees showing comb jellies as the sister of all the other
animals are artifactual

Here are some excerpts from their abstract:

> Increased model adequacy is associated with significant topological changes in support of Porifera-sister.

> Because adequate modeling of the evolutionary process that generated the data is fundamental to recovering an accurate phylogeny [16–20], our results strongly support sponges as the sister group of all other animals and provide further evidence that Ctenophora-sister represents a tree reconstruction artifact.

On social media and in comments to the press the tone of their claims have been even stronger, where they have declared this manuscript to be the last word in the debate and the question to be resolved. Take, for example, the following quote by one of the study co-authors from the Guardian article *[Evolution row ends as scientists declare sponges to be sister of all other animals](https://www.theguardian.com/science/2017/nov/30/evolution-row-ends-as-scientists-declare-sponges-to-be-sister-of-all-animals?CMP=share_btn_tw)*:

> While [co-author] Pisani believes proponents of the comb jellies might continue the debate, he says the wrangle is drawing to a close. “From my perspective, yes, this is the last word,” he said.

Strong claims of course require strong evidence, so let's take a look at the evidence presented in their manuscript.

```{r load_data}

library( tidyverse )
library( magrittr )
library( scales )

D = read.table( "tidy_table_3.tsv", header=TRUE, stringsAsFactors=FALSE  )

D$Model = factor( D$Model, c("GTR+G", "CAT-GTR+G" ) )

D$Dataset = sub( "D20", "", D$Dataset )
D$Dataset = sub( "Chang", "Chang-Holo", D$Dataset )
D$Dataset = sub( "Opistho", "Opis", D$Dataset )
D$Dataset = sub( "Choano", "Cho", D$Dataset )
D$Dataset = factor( D$Dataset, c( "Whelan-Cho", "Whelan-Holo", "Whelan-Opis", "Chang-Holo" ) )

D$Recoding[ D$Recoding == "KGB-6" ] = "K"
D$Recoding[ D$Recoding == "Dayhoff-6" ] = "D"
D$Recoding[ D$Recoding == "SR-6" ] = "S"
D$Recoding = factor( D$Recoding, c("None", "K", "D", "S" ) )

pp_thresh = 0.95
D$Result = "Unresolved"
D$Result[ D$Porifera_sister >= pp_thresh ] = "Porifera-sister" 
D$Result[ D$Ctenophora_sister >= pp_thresh ] = "Ctenophora-sister" 
D$Result = factor( D$Result )

D$Convergence = "bad"
D$Convergence[ D$Maxdiff <= 1 ] = "possible"
D$Convergence[ D$Maxdiff <= 0.3 ] = "acceptable"
D$Convergence[ D$Maxdiff <= 0.1 ] = "good"
D$Convergence = factor( D$Convergence, c( "good", "acceptable", "possible", "bad" ) )

cteno_color = "#F8766D"
poro_color = "#00BA38"

```

## The results of Feuda *et al.* 2017

The central findings of Feuda *et al.* 2017 are presented in their Table 3. It is reproduced below.

![table 3 feuda](figures/table_3_feuda.png)

*Table 3 from Feuda et al. 2017.*

I reorganized the data in this table and color coded the cells to get a better view the general pattern of support for Porifera-sister. This reorganized table is presented below as Figure 2.

```{r result_pp}
D %>%
ggplot( aes( Recoding, Dataset ) ) + 
	geom_tile(aes(fill = Porifera_sister)) + 
	facet_wrap(~ Model) +
	scale_fill_gradient(low = cteno_color, high = poro_color, limits=c(0, 1)) + 
	geom_text(aes(label = Porifera_sister)) +
	ggtitle("Porifera-sister Support") +
	theme_classic()
```

*Figure 2 - Posterior probabilities for Porifera-sister. Recodings are: None- original amino acid data, K - KGB-6, D - Dayhoff-6, and S - SR-6.*

I next color coded the results by significance, assessed as a posterior probability &ge; `r pp_thresh`. This is a low threshold, most Bayesian phylogenetic studies apply a more stringent (higher) threshold to assess significance. The significance of the results is presented below in Figure 3.

```{r result_sig, fig.cap="Phylogenetic results of Feuda. These results were parsed from thier Table 3."}

D %>%
ggplot( aes( Recoding, Dataset ) ) + 
	geom_tile( aes( fill = Result ) ) + 
	facet_wrap( ~ Model ) +
	ggtitle("Ctenophore-sister Support") +
	theme_classic()

```

*Figure 3 - Significance of posterior probabilities for Porifera-sister assessed as posterior probability &ge; `r pp_thresh`. *

There are a few striking patterns in these results, which I explore below.

### Model

All analyses under GTR+G (Figure 3, left panel) recover significant support for Ctenophora-sister, and some analyses (Figure 3, right panel) under CAT-GTR+G recover Porifera-sister.

Feuda *et al.* 2017 examine model fit and adequacy and come to the conclusion that the CAT-GTR+G model is the best available model for the data. Let's therefore only consider the more complex CAT-GTR+G model further.

The results are quite variable under the CAT-GTR+G model (Figure 3, right panel). Let's break this down:

- Both full matrices support Ctenophora-sister under the CAT-GTR+G model (Figure 3, right panel, top two left cells). 
- Most analyses under the CAT-GTR+G model do not provide significant support for either Ctenophora-sister or Porifera-sister (Figure 3, right panel, 10 blue cells)
- A subset of matrices with reduced species sampling and recoded data provide significant support for Porifera-sister (Figure 3, right panel, 4 green cells)

There is significant support for Porifera-sister in only one quarter of CAT-GTR+G analyses.

### Data reduction

The different results within the 16 CAT-GTR+G analyses are due to different approaches to data reduction: (1) recoding data in both the Whelan and Chang matrix, and (2) removing outgroups from the Whelan matrix.

While the full amino acid matrix of Chang (Figure 3, right panel, top left red cell) provides significant support for Ctenophora-sister, the recoded analyses are unresolved (Figure 3, right panel, remainder of top row) and none provide significant support for Porifera-sister.

Authors of Feuda *et al.* 2017, in that manuscript and others, have strongly argued that broader outgroup sampling has a *negative* impact on this particular phylogenetic question. This is counter to the common understanding in the field that broad outgroup sampling is critical to rooting problems like this one. Given this assertion, it is interesting that reducing outgroup sampling in the Whelan matrix (Figure 3, right panel, left row bottom two cells) does not provide significant support for Porifera-sister.

The strong conclusions the authors made based largely on Table 3 are therefore rest entirely on recoding the Whelan matrix and applying the CAT-GTR+G model. Even then, less than half (4/9) of these recoded Whelan CAT-GTR+G analyses (Figure 3, right panel, bottom right 9 cells) provide significant support for Porifera-sister.

### Reporting results

Some of these results in Table 3 were clearly described by Feuda *et al.* 2017, and others were not addressed. There are, however, also some inconsistencies between the results and the description of the results in the text. The overall impact is that the text presents much stronger support for Porifera-sister than is actually present in the results.

For example:

> Porifera-sister was favored under all recoding strategies.

This is not the case. Only 4 of the twelve analyses of recoded data presented in Table 3 show significant support for Porifera-sister (Figure 3). Some recoded analyses have a posterior probability < 0.5 for Porifera-sister.

> Data recoding reduced the impact of heterogeneity on our phylogenetic analyses (see also [7]), and this lent support to Porifera-sister. Notably, this result emerged independently of the outgroups included.

This is not the case. The recoded Whelan CAT-GTR+G analyses are sensitive to outgroup sampling, both in magnitude of posterior probability (Figure 2, right panel, bottom right 9 cells) and significance of posterior probability (Figure 3, right panel, bottom right 9 cells).

## Moving beyond the last word

Feuda *et al.* 2017 was not the last word in resolving the root of the animal tree. So what next?

One of the most critical steps to take is to change the way we talk about the problem. In particular, insignificant phylogenetic results should not be presented as definitive resolutions of tough problems. This should always apply, but especially so in the case of highly contested questions of such broad interest. Making strong statements about insignificant results gives the impression that people with different interpretations of the available results are much further apart than they actually are.

Overstating results has given the entirely wrong impression to the broader community of where the Ctenophora-sister and Porifera-sister hypotheses stand. It is not the case that different groups of investigators are doing the same thing and getting radically different results. They are doing different things, getting slightly different results, and coming to very different conclusions. Making strong statements about insignificant results eventually gives the impression that phylogenetic methods are broken. They are not. How we talk about phylogenies is broken.

When I look at the entirety of the impressive work done by multiple very talented groups on testing Ctenophora-sister and Porifera-sister, I come to the conclusion that at this time we cannot reject either hypothesis. I just don't find the sensitivity to species sampling, gene sampling, model selection, and now character coding conclusive. It isn't that I don't think it is impossible to come to a well-supported result if support for a phylogenetic hypothesis is sensitive to any of these factors, I just don't think we understand the sensitivity of this problem well enough to come to strong conclusions.

I also find the interactions between these factors concerning. I think that the rigorously blocked approach that Feuda *et al.* 2017 very exciting and the way of the future. It can provide a clear picture of the support landscape as it depends on different factors and their interactions. We are just looking at too small a part of the sensitivity landscape at too coarse a scale to understand the problem well enough yet.

I am optimistic that there will be significant progress on this problem in coming years. Analysis methods and tools are rapidly improving. New high-quality genome data will soon be available for many more species that are highly relevant to this question. For example, just a few days after Feuda *et al.* 2017 was published, [Belahbib et al.](https://www.biorxiv.org/content/early/2017/12/04/228452) posted a preprint presenting two new sponge genome sequences and analyses directly relevant to interpreting the biological significance of Ctenophora-sister and Porifera-sister. I look forward to to seeing how further advances in analyses and sampling impact tests of Ctenophora-sister and Porifera-sister and their impact on our understanding of animal evolution.

### 2017-12-10 readme.rmd
light edits

@@
 
-In a recent paper, [Feuda *et al.* 2017](https://doi.org/10.1016/j.cub.2017.11.008) claim to present strong significant support for Porifera-sister and to have resolved the question at last.
+In a recent paper, [Feuda *et al.* 2017](https://doi.org/10.1016/j.cub.2017.11.008) claim to present strong, significant support for Porifera-sister and to have resolved the question at last.
 Feuda *et al.* present a well-blocked set of phylogenetic analyses that examine support for Ctenophora-sister and Porifera-sister under a variety of conditions.
@@
 
-Here are the highlights that accompany their paper:
+Here are a couple of the highlights that accompany their paper:
 
-> Highlights
->
-> This is the first comparison of model adequacy in non- bilaterian phylogenomics
->
-> Animal relationships depend on the ability of the models to describe the data
->
 > Sponges are the sister group of the remaining animals
@@
 
-I reorganized the data in this table and color coded the cells to get a better view the general pattern of support for Porifera-sister. This reorganized table is presented below as Figure 2.
+I reorganized the data in this table and color coded the cells to get a better view of the general pattern of support for Porifera-sister. This reorganized table is presented below as Figure 2.
 
@@
 
-*Figure 2 - Posterior probabilities for Porifera-sister. Recodings are: None- original amino acid data, K - KGB-6, D - Dayhoff-6, and S - SR-6.*
+*Figure 2 - Posterior probabilities for Porifera-sister. Recodings are: None- original amino acid data, K - KGB-6, D - Dayhoff-6, and SR-6.*
 
@@
 
-The strong conclusions the authors made based largely on Table 3 are therefore rest entirely on recoding the Whelan matrix and applying the CAT-GTR+G model.
+The strong conclusions the authors made based largely on Table 3 therefore rest entirely on recoding the Whelan matrix and applying the CAT-GTR+G model.
 Even then, less than half (4/9) of these recoded Whelan CAT-GTR+G analyses (Figure 3, right panel, bottom right 9 cells) provide significant support for Porifera-sister.
@@
 New high-quality genome data will soon be available for many more species that are highly relevant to this question.
-For example, just a few days after Feuda *et al.* 2017 was published, [Belahbib et al.](https://www.biorxiv.org/content/early/2017/12/04/228452) posted a preprint presenting two new sponge genome sequences and analyses directly relevant to interpreting the biological significance of Ctenophora-sister and Porifera-sister.
+For example, just a few days after Feuda *et al.* 2017 was published, [Belahbib *et al.*](https://www.biorxiv.org/content/early/2017/12/04/228452) posted a preprint presenting two new sponge genome sequences and analyses directly relevant to interpreting the biological significance of Ctenophora-sister and Porifera-sister.
 I look forward to to seeing how further advances in analyses and sampling impact tests of Ctenophora-sister and Porifera-sister and their impact on our understanding of animal evolution.

### 2017-12-10 readme.rmd
increased figure resolution

@@
 	warning = FALSE,
-	echo=FALSE
+	echo=FALSE,
+	dpi=300
 	)
@@
 
-- Both full matrices support Ctenophora-sister under the CAT-GTR+G model (Figure 3, right panel, top two left cells). 
+- Both full matrices support Ctenophora-sister under the CAT-GTR+G model (Figure 3, right panel, left column, top two cells). 
 - Most analyses under the CAT-GTR+G model do not provide significant support for either Ctenophora-sister or Porifera-sister (Figure 3, right panel, 10 blue cells)

### 2017-12-10 readme.rmd
more on recoding

@@
 
-*Figure 2 - Posterior probabilities for Porifera-sister. Recodings are: None- original amino acid data, K - KGB-6, D - Dayhoff-6, and SR-6.*
+*Figure 2 - Posterior probabilities for Porifera-sister. Recodings are: None- original amino acid data, K - KGB-6, D - Dayhoff-6, and S- SR-6.*
 
@@
 
-Authors of Feuda *et al.* 2017, in that manuscript and others, have strongly argued that broader outgroup sampling has a *negative* impact on this particular phylogenetic question.
+Authors of Feuda *et al.* 2017, in that manuscript and others (*e.g.* [Pick *et al.* 2010](https://doi.org/10.1093/molbev/msq089), have strongly argued that broader outgroup sampling has a *negative* impact on this particular phylogenetic question.
 This is counter to the common understanding in the field that broad outgroup sampling is critical to rooting problems like this one.
@@
 The strong conclusions the authors made based largely on Table 3 therefore rest entirely on recoding the Whelan matrix and applying the CAT-GTR+G model.
-Even then, less than half (4/9) of these recoded Whelan CAT-GTR+G analyses (Figure 3, right panel, bottom right 9 cells) provide significant support for Porifera-sister.
+These results are central to the paper since they include the only results with significant support for Porifera-sister.
+The authors also indicate that these recodings do a good job addressing issues they are concerned with in other analyses:
+
+> All recoded datasets obtained lower Z scores than the original amino acid datasets for both PPA-MAX and PPA- MEAN (Tables 1 and S1), indicating that recoding is effective at reducing lineage-specific compositional heterogeneity.
+
+Despite the author claim that this method addresses key technical problems, less than half (4/9) of these recoded Whelan CAT-GTR+G analyses (Figure 3, right panel, bottom right 9 cells) provide significant support for Porifera-sister.
+The results are still sensitive to the recoding strategy used and outgroup sampling.
+It also calls into question the robustness of results under recoding strategies.
+For example, none of the KGB-6 recoded analyses provide significant support for Porifera-sister, despite having improved adequacy similar to that of other recodings and relying on very similar data reduction transforms.
+
+Though the authors interpreted these recoded analyses as showing conclusive support for Porifera-sister, closer examination of the results instead indicates that improved method adequacy does not always provide stronger support for Porifera-sister and variation in support remains largely unexplained.
 
@@
 Some of these results in Table 3 were clearly described by Feuda *et al.* 2017, and others were not addressed.
-There are, however, also some inconsistencies between the results and the description of the results in the text.
+There are, however, also some inconsistencies between the presented results and the description of the results in the text.
 The overall impact is that the text presents much stronger support for Porifera-sister than is actually present in the results.
@@
 Making strong statements about insignificant results gives the impression that people with different interpretations of the available results are much further apart than they actually are.
+Making strong statements about insignificant results eventually gives the impression that phylogenetic methods are broken.
+They are not.
+How we talk about phylogenies is broken.
 
@@
 It is not the case that different groups of investigators are doing the same thing and getting radically different results.
-They are doing different things, getting slightly different results, and coming to very different conclusions.
-Making strong statements about insignificant results eventually gives the impression that phylogenetic methods are broken.
-They are not.
-How we talk about phylogenies is broken.
+They are looking at similar results that consistently show sensitivity to multiple factors, and drawing very different conclusions about the implications of that sensitivity.
 

### 2017-12-10 readme.rmd
added support landscape

@@
 
+## Using sensitivity to understand the support landscape
+
+The sensitivity of phylogenetic analyses to taxon sampling, gene sampling, and analysis methods is often viewed as troublesome result.
+It provides very important insight, though, into the methods we use and the biological questions we apply them to.
+After ten years of intensive study, the placement of sponges and ctenophores have become one of the most thoroughly explored difficult phylogenetic problems.
+Examining the sensitivity of results will help us advance this particular question, and may also provide insight into how to address other hard phylogenetic questions.
+
+Feuda *et al.* explain why they interpret the sensitivity they and others present as support for Porifera-sister in a few passages, including:
+
+> The relationships at the root of the animal tree depend on model adequacy, with Ctenophora-sister emerging more prominently when the data are modeled less adequately and Porifera-sister being better supported when the data are more adequately modeled (Figure 1).
+As phylogenetic methods are less likely to recover artifactual topologies when the models of evolution more adequately describe the data [16–20], we conclude that Porifera-sister better represents the phylogenetic signal in the data.
+
+> This model-dependent outcome has been interpreted as reflecting a fundamental lack of robustness [59], and it has been suggested that the relationships at the root of the animal tree might be impossible to reconstruct based on current amino acid datasets [2, 4].
+However, sensitivity to the model (and to other analytical factors) does not imply an impossibility to decide which solution is most likely to be correct...
+the varied outcomes of phylogenetic analyses of animal relationships show a clear pattern, with Ctenophora-sister being systematically associated with the use of inadequate models and/or taxon-sampling schemes that are most likely to exacerbate systematic errors.
+
+It is certainly true that model sensitivity alone does not mean we can reject a given phylogenetic hypothesis.
+There are an infinite number of possible models, and it is essentially certain that for any given phylogenetic hypothesis there is a model out there somewhere that will reject it.
+The question is how we use variation in support to put each result in a broader context that allows us to interpret it.
+
+We can think of this variation as a support landscape - a multidimensional space that shows how support varies across different gene sampling, taxon sampling, and analysis methods.
+It is only ever possible to probe a finite region of this landscape, but its overall shape has radical implications for how we interpret a particular set of results.
+This means that when investigating sensitivity it is necessary to be explicit not only about the results we get in a given study, but how we are thinking about regions of the support landscape that are adjacent to those we examine.
+
+The excerpts from Feuda *et al.* suggest that the authors implicitly envision the broader support landscape to be a monotonic surface (Figure 4a).
+They see believe that the small part of the support landscape they probed (Figure 4, inset boxes) indicates a broader pattern that they interpret as strong support for Porifera-sister.
+This is because, although most analyses they did support Ctenophora-sister there are some analyses on the right edge of the sampled space that recover significant support for Porifera-sister.
+
+![overview](figures/figure_4.png)
+*Figure 4 - Hypothetical support landscapes.
+The inset box shows the region of the landscape explored by Figures 2-3.
+The horizontal axis represents variation in analysis methods, such as model of molecular evolution and recoding of data.
+The vertical axis represents variation in sampling of genes and species.
+(a) A Monotonic Support Landscape.
+(b) A Sponge-Island Support Landscape, which would indicate that sponges support is only obtained in a very small portion of the support landscape.
+(c) A Hilly Support Landscape.*
+
+Other support landscapes would have very different implications for the results they obtained.
+A small fraction of analyses may have supported Porifera-sister because they happened to sample a small island of support (Figure 4b) in a broader landscape that overall his patterns of support consistent with Ctenophora-sister.
+Alternatively, the small island of support for Porifera-sister may indicate a transition from a plateau of support for Ctenophora sister to a hilly support landscape to the east.
+It is unlikely that we would conclusively resolve the placement of ctenophores and sponges if this broader hilly pattern were found to be the case.
+
+The work of Feuda *et al.* 2017 provides an enticing peek at a portion of this landscape, but without broader context still fails to allow us to come to strong conclusions.
+Exploring additional sensitivity will help us understand this bigger picture.
+In this sense, sensitivity is an asset rather than a liability.
+
 ## Moving beyond the last word
@@
 This should always apply, but especially so in the case of highly contested questions of such broad interest.
-Making strong statements about insignificant results gives the impression that people with different interpretations of the available results are much further apart than they actually are.
-Making strong statements about insignificant results eventually gives the impression that phylogenetic methods are broken.
+Making strong statements about insignificant results gives the impression that phylogenetic methods are broken as a question seems to bounce from one extreme answer to the other.
 They are not.
@@
 I just don't find the sensitivity to species sampling, gene sampling, model selection, and now character coding conclusive.
-It isn't that I don't think it is impossible to come to a well-supported result if support for a phylogenetic hypothesis is sensitive to any of these factors, I just don't think we understand the sensitivity of this problem well enough to come to strong conclusions.
+This is because I think we need a broader perspective of the support landscape to put the interesting results we already have into context.
+It isn't that I don't think it is impossible to come to a well-supported result if support for a phylogenetic hypothesis is sensitive to any of these factors, I just don't think we understand the sensitivity of this problem well enough to come to strong conclusions at this time.
 
@@
 It can provide a clear picture of the support landscape as it depends on different factors and their interactions.
-We are just looking at too small a part of the sensitivity landscape at too coarse a scale to understand the problem well enough yet.
+We need to expand this approach.
 
@@
 For example, just a few days after Feuda *et al.* 2017 was published, [Belahbib *et al.*](https://www.biorxiv.org/content/early/2017/12/04/228452) posted a preprint presenting two new sponge genome sequences and analyses directly relevant to interpreting the biological significance of Ctenophora-sister and Porifera-sister.
-I look forward to to seeing how further advances in analyses and sampling impact tests of Ctenophora-sister and Porifera-sister and their impact on our understanding of animal evolution.
+I look forward to to seeing how further advances in analyses and sampling push tests of Ctenophora-sister and Porifera-sister into new areas of the support landscape and allow us to chart its broader features.
+And, of course, I am excited about the impact this work will have on our understanding of animal evolution.

### 2017-12-10 readme.rmd
added to last section

@@
 I just don't find the sensitivity to species sampling, gene sampling, model selection, and now character coding conclusive.
-This is because I think we need a broader perspective of the support landscape to put the interesting results we already have into context.
-It isn't that I don't think it is impossible to come to a well-supported result if support for a phylogenetic hypothesis is sensitive to any of these factors, I just don't think we understand the sensitivity of this problem well enough to come to strong conclusions at this time.
+I also find the interactions between these factors concerning.
+We need a broader perspective of the support landscape to put the interesting results we already have into context before we can understand the implications of sensitivity for these hypotheses.
 
-I also find the interactions between these factors concerning.
+GTR+G and CAT-GTR+G are very similar models that were implemented in carefully controlled ways by Feuda *et al.* 2017, yet provide very different results.
+What will happen when we consider other similar, adjacent models?
+It isn't surprising when highly different models give different results, but when adding a single set of parameters gives a radically different result we should be skeptical that we have the final word in the problem under either model.
+The fact that the authors claim that "recoding is effective at reducing lineage-specific compositional heterogeneity", yet still get highly inconsistent results in different recoded analyses is also problematic.
+It is also interesting to note that authors of Feuda *et al.* declared another set of analyses ([Pisani *et al.* 2015](http://dx.doi.org/10.1073/pnas.1518127112)) to be the [nail in the coffin of Ctenophora sister](https://twitter.com/gwoerhe/status/671636958926499841) even though it did not use the recoding strategy that they now claim to be essential to testing these hypotheses.
+Such strong claims manufacture controversy where there is none and, in combination with moving goal posts, give the impression that the only consistent criterion they apply for detecting problems with analyses is finding that they support Ctenophora-sister.
+
 I think that the rigorously blocked approach that Feuda *et al.* 2017 very exciting and the way of the future.
@@
 And, of course, I am excited about the impact this work will have on our understanding of animal evolution.
+
+Porifera may very well be the sister group to all other animals, but declaring them to be sister based on inconsistent and inconclusive results will make it harder to find out if they are.
+It discourages the very work that still needs to be done to test Porifera-sister and Ctenophora-sister.
+It also alienates so many people who want to know the answer (and to make progress on other hard phylogenetic problems) because it makes it seem like phylogenetic methods give radically different results from one day to the next.

### 2017-12-10 readme.rmd
typos

@@
 The excerpts from Feuda *et al.* suggest that the authors implicitly envision the broader support landscape to be a monotonic surface (Figure 4a).
-They see believe that the small part of the support landscape they probed (Figure 4, inset boxes) indicates a broader pattern that they interpret as strong support for Porifera-sister.
+They believe that the small part of the support landscape they probed (Figure 4, inset boxes) indicates a broader pattern that they interpret as strong support for Porifera-sister.
 This is because, although most analyses they did support Ctenophora-sister there are some analyses on the right edge of the sampled space that recover significant support for Porifera-sister.
@@
 Other support landscapes would have very different implications for the results they obtained.
-A small fraction of analyses may have supported Porifera-sister because they happened to sample a small island of support (Figure 4b) in a broader landscape that overall his patterns of support consistent with Ctenophora-sister.
+A small fraction of analyses may have supported Porifera-sister because they happened to sample a small island of support (Figure 4b) in a broader landscape that overall has patterns of support consistent with Ctenophora-sister.
 Alternatively, the small island of support for Porifera-sister may indicate a transition from a plateau of support for Ctenophora sister to a hilly support landscape to the east.

### 2017-12-10 readme.rmd
added press release quote

@@
 > While [co-author] Pisani believes proponents of the comb jellies might continue the debate, he says the wrangle is drawing to a close. “From my perspective, yes, this is the last word,” he said.
+
+And from a press release *[In the beginning there was the sponge](http://www.en.uni-muenchen.de/news/newsarchiv/2017/woerheide_schwammanalysen.html)* describing their new paper:
+
+> ...
+their results are unequivocal: The sponges form the earliest branch on the animals' family tree...
+"Our results confirm the classical assumptions concerning early animal evolution, and should help to put an end to the recent controversy over the origin of multicellular animals," Wörheide concludes.
 

### 2017-12-11 readme.rmd
word smithing

@@
 
-## Using sensitivity to understand the support landscape
+## Sensitivity and the support landscape
 
@@
 
-Feuda *et al.* explain why they interpret the sensitivity they and others present as support for Porifera-sister in a few passages, including:
+Feuda *et al.* explain why they interpret their results as evidence for Porifera-sister, even though support is sensitive to multiple factors, as follows:
 
@@
 One of the most critical steps to take is to change the way we talk about the problem.
-In particular, insignificant phylogenetic results should not be presented as definitive resolutions of tough problems.
+In particular, non-significant phylogenetic results should not be presented as definitive resolutions of tough problems.
 This should always apply, but especially so in the case of highly contested questions of such broad interest.
-Making strong statements about insignificant results gives the impression that phylogenetic methods are broken as a question seems to bounce from one extreme answer to the other.
+Making strong statements about non-significant results gives the impression that phylogenetic methods are broken as a question seems to bounce from one extreme answer to the other.
 They are not.

### 2017-12-11 readme.rmd
made figure cells square, landscapes curved

@@
 there is an internal branch in the animal phylogeny that separates sponges from all other animals (shown in green).
-Other relationships that are not part of these hypotheses are shown as an unresolved polytomy.
+Other relationships that are not part of these hypotheses are shown as unresolved polytomies.
 Figure adapted from [Dunn et al.
@@
 - Character recoding.
-They recoded amino acids with similar chemical properties as the same character using three different schemes (KGB-6, Dayhoff-6, SR-6).
+They recoded amino acids that are often exchanged for each other in the course of evolution as the same character using three different schemes (KGB-6, Dayhoff-6, SR-6).
+In each case, the 20 amino acids are reduced to 6 categories, with different amino acids in each category for each scheme (see their Table 2 for details).
 This recoding is a form of data reduction - rather than remove species or sites, it removes dynamic range within each cell of the matrix.
-The motivation of this recoding is to minimize the impacts of compositional bias, since different lineages and genes may tend to use different amino acids with similar properties.
+The motivation of this recoding is to minimize the impacts of compositional bias, since different lineages and genes may tend to use different amino acids.
 
@@
 
-```{r result_pp}
+```{r result_pp, fig.width=6.5, fig.height=3.1}
 D %>%
@@
 
-```{r result_sig, fig.cap="Phylogenetic results of Feuda. These results were parsed from thier Table 3."}
+```{r result_sig, fig.width=6.5, fig.height=3.1}
 
@@
 
-Authors of Feuda *et al.* 2017, in that manuscript and others (*e.g.* [Pick *et al.* 2010](https://doi.org/10.1093/molbev/msq089), have strongly argued that broader outgroup sampling has a *negative* impact on this particular phylogenetic question.
+Authors of Feuda *et al.* 2017, in that manuscript and others (*e.g.* [Pick *et al.* 2010](https://doi.org/10.1093/molbev/msq089)), have strongly argued that broader outgroup sampling has a *negative* impact on this particular phylogenetic question.
 This is counter to the common understanding in the field that broad outgroup sampling is critical to rooting problems like this one.

### 2017-12-11 readme.rmd
fig formatting

@@
 	facet_wrap( ~ Model ) +
-	ggtitle("Ctenophore-sister Support") +
+	ggtitle("Animal Root Support") +
 	theme_classic()

### 2017-12-11 readme.rmd
added dash

@@
 The fact that the authors claim that "recoding is effective at reducing lineage-specific compositional heterogeneity", yet still get highly inconsistent results in different recoded analyses is also problematic.
-It is also interesting to note that authors of Feuda *et al.* declared another set of analyses ([Pisani *et al.* 2015](http://dx.doi.org/10.1073/pnas.1518127112)) to be the [nail in the coffin of Ctenophora sister](https://twitter.com/gwoerhe/status/671636958926499841) even though it did not use the recoding strategy that they now claim to be essential to testing these hypotheses.
+It is also interesting to note that authors of Feuda *et al.* declared another set of analyses ([Pisani *et al.* 2015](http://dx.doi.org/10.1073/pnas.1518127112)) to be the [nail in the coffin of Ctenophora-sister](https://twitter.com/gwoerhe/status/671636958926499841) even though it did not use the recoding strategy that they now claim to be essential to testing these hypotheses.
 Such strong claims manufacture controversy where there is none and, in combination with moving goal posts, give the impression that the only consistent criterion they apply for detecting problems with analyses is finding that they support Ctenophora-sister.

### 2017-12-11 readme.rmd
minor edits throughout

@@
 This is counter to the common understanding in the field that broad outgroup sampling is critical to rooting problems like this one.
-Given this assertion, it is interesting that reducing outgroup sampling in the Whelan matrix (Figure 3, right panel, left row bottom two cells) does not provide significant support for Porifera-sister.
+Given this assertion, it is interesting that reducing outgroup sampling in the Whelan amino acid matrix (Figure 3, right panel, left column, bottom two cells) does not provide significant support for Porifera-sister.
 
@@
 The results are still sensitive to the recoding strategy used and outgroup sampling.
-It also calls into question the robustness of results under recoding strategies.
 For example, none of the KGB-6 recoded analyses provide significant support for Porifera-sister, despite having improved adequacy similar to that of other recodings and relying on very similar data reduction transforms.
+This calls into question the robustness of results under recoding strategies.
 
@@
 
-The sensitivity of phylogenetic analyses to taxon sampling, gene sampling, and analysis methods is often viewed as troublesome result.
+Sensitivity of phylogenetic analyses to taxon sampling, gene sampling, and analysis methods is often viewed as a troublesome result.
 It provides very important insight, though, into the methods we use and the biological questions we apply them to.
@@
 
-We can think of this variation as a support landscape - a multidimensional space that shows how support varies across different gene sampling, taxon sampling, and analysis methods.
+We can think of this variation as a support landscape (Figure 4) - a multidimensional space that shows how support varies across different gene sampling, taxon sampling, and analysis methods.
 It is only ever possible to probe a finite region of this landscape, but its overall shape has radical implications for how we interpret a particular set of results.
@@
 They believe that the small part of the support landscape they probed (Figure 4, inset boxes) indicates a broader pattern that they interpret as strong support for Porifera-sister.
-This is because, although most analyses they did support Ctenophora-sister there are some analyses on the right edge of the sampled space that recover significant support for Porifera-sister.
+This is because, although most of their analyses support Ctenophora-sister there are some analyses on the right edge of the sampled space that recover significant support for Porifera-sister.
 
@@
 
-One of the most critical steps to take is to change the way we talk about the problem.
+First, it is critical to change the way we talk about the problem.
 In particular, non-significant phylogenetic results should not be presented as definitive resolutions of tough problems.
@@
 Making strong statements about non-significant results gives the impression that phylogenetic methods are broken as a question seems to bounce from one extreme answer to the other.
-They are not.
+Phylogenetic methods are not broken.
 How we talk about phylogenies is broken.
@@
 It discourages the very work that still needs to be done to test Porifera-sister and Ctenophora-sister.
-It also alienates so many people who want to know the answer (and to make progress on other hard phylogenetic problems) because it makes it seem like phylogenetic methods give radically different results from one day to the next.
+It also alienates so many people who want to know the answer (and to make progress on other hard phylogenetic problems).

### 2017-12-12 readme.rmd
minor wording

@@
 I next color coded the results by significance, assessed as a posterior probability &ge; `r pp_thresh`.
-This is a low threshold, most Bayesian phylogenetic studies apply a more stringent (higher) threshold to assess significance.
+This is a low threshold.
+Most Bayesian phylogenetic studies apply a more stringent (higher) threshold to assess significance.
 The significance of the results is presented below in Figure 3.
@@
 
-The work of Feuda *et al.* 2017 provides an enticing peek at a portion of this landscape, but without broader context still fails to allow us to come to strong conclusions.
+The work of Feuda *et al.* 2017 provides an enticing glimpse at a portion of this landscape, but without broader context still fails to allow us to come to strong conclusions.
 Exploring additional sensitivity will help us understand this bigger picture.
@@
 
-I think that the rigorously blocked approach that Feuda *et al.* 2017 very exciting and the way of the future.
+I think that the rigorously blocked approach of Feuda *et al.* 2017 is very exciting and the way of the future.
 It can provide a clear picture of the support landscape as it depends on different factors and their interactions.

### 2017-12-12 readme.rmd
added details on focal analyses

@@
 
+D$Maxdiff = round( D$Maxdiff, 2 )
+
 cteno_color = "#F8766D"
 poro_color = "#00BA38"
+unresolved_color = "#6F94CD"
 
@@
 
+gg_sig = 
 	D %>%
@@
 			theme_classic()
+
+ggsave( "figures/figure_3.pdf", gg_sig, device=pdf, width=6.5, height=3.1 )
+
+gg_sig
 
@@
 
-## Sensitivity and the support landscape
+## Sensitivity within the presented results
+
+The primary explanation Feuda *et al.* 2017 provide for accepting some results and rejecting others is their posterior predictive (PP) analyses of model adequacy.
+They calculated PP for only the Whelan-Opisthokont and Chang matrices (Figures 2-3, top two rows).
+Based on these scores they rejected the results of the GTR-G analyses (Figures 2-3, left pane) and analyses that had not been recoded.
+This left six cells that are the focus of their primary conclusions (Figures 2-3, right panel, top two rows, right 3 cells).
+
+```{r result_sub}
+
+D_sub = 
+	D %>%
+	rename( Porifera_sister_posterior = Porifera_sister ) %>%
+	filter(Model=="CAT-GTR+G") %>% 
+	filter( (Dataset=="Whelan-Opis") | (Dataset=="Chang-Holo") ) %>% 
+	filter( Recoding != "None" ) %>%
+	select (-c(Ctenophora_sister, Burnin, effsize, rel_diff)) %>% 
+	gather( var, val, Porifera_sister_posterior:PPA_MEAN ) %>% 
+	mutate( var = factor( var, levels = c("Porifera_sister_posterior", "Cycles", "Maxdiff", "PPA_DIV", "PPA_MAX", "PPA_CONV", "PPA_VAR", "PPA_MEAN")))
+
+gg_sub = 
+	D_sub %>%
+		ggplot( aes( Recoding, Dataset ) ) + 
+			geom_tile( aes( fill = Result ) ) + 
+			facet_wrap( ~ var, ncol = 2 ) +
+			geom_text(aes(label = val)) +
+			scale_fill_manual( values=c( poro_color,unresolved_color ) ) +
+			ggtitle("Variation between analyses") +
+			theme_classic()
+
+ggsave( "figures/figure_4_raw.pdf", gg_sub, device=pdf, width=6.5, height=6 )
+
+```
+
+![figure 4](figures/figure_4.png)
+
+*Figure 4 - Detailed examination of the analysis subset (indicated in upper right pane) that is the focus of the conclusions by Feuda et al.
+These are the six recoded CAT-GTR+G analyses with posterior predictive (PP) scores.
+The presented statistics for these six cells are posterior probability of Poifera-sister, Maxdiff (with lower scores indicating better convergence), and five PP scores (where lower absolute value indicates better model adequacy).
+Only one analysis provides significant support for Porifera-sister.
+It isn't the most adequate by any of the PP scores, and showed the poorest convergence according to Maxdiff.
+Recodings are: None- K - KGB-6, D - Dayhoff-6, and S- SR-6.*
+
+A variety of statistics from Feuda *et al.* 2017 for these six cells are presented in Figure 4.
+The first thing to note is that only one of these six analyses provides significant support for Porifera-sister.
+This is for the SR-6 recoded Whelan-Opisthokont matrix, shown in green in Figure 4.
+Given the similarity of these six analyses, the fact that only one supports Porifera-sister calls into serious question the robustness of support for Porifera-sister.
+
+The other statistics help put this single significant result into context, and they could be used to provide explicit criteria for favoring the one significant result over the five insignificant results.
+Feuda *et al.* present PP as the primary criterion we should consider.
+The one analysis with significant support for Porifera-sister does not have the best (lowest absolute value) PP score for any of the five PP statistics considered (Figure 4).
+By their own criterion of model adequacy, we should favor the unresolved analyses.
+
+In fact, the only statistic that stands out for the significant analysis is the high Maxdiff score, which indicates it is the most poorly converged of these six focal analyses.
+
+## Sensitivity and the broader support landscape
 
@@
 
-We can think of this variation as a support landscape (Figure 4) - a multidimensional space that shows how support varies across different gene sampling, taxon sampling, and analysis methods.
+We can think of this variation as a support landscape (Figure 5) - a multidimensional space that shows how support varies across different gene sampling, taxon sampling, and analysis methods.
 It is only ever possible to probe a finite region of this landscape, but its overall shape has radical implications for how we interpret a particular set of results.
@@
 
-The excerpts from Feuda *et al.* suggest that the authors implicitly envision the broader support landscape to be a monotonic surface (Figure 4a).
-They believe that the small part of the support landscape they probed (Figure 4, inset boxes) indicates a broader pattern that they interpret as strong support for Porifera-sister.
+The excerpts from Feuda *et al.* suggest that the authors implicitly envision the broader support landscape to be a monotonic surface (Figure 5a).
+They believe that the small part of the support landscape they probed (Figure 5, inset boxes) indicates a broader pattern that they interpret as strong support for Porifera-sister.
 This is because, although most of their analyses support Ctenophora-sister there are some analyses on the right edge of the sampled space that recover significant support for Porifera-sister.
 
-![overview](figures/figure_4.png)
-*Figure 4 - Hypothetical support landscapes.
+![overview](figures/figure_5.png)
+*Figure 5 - Hypothetical support landscapes.
 The inset box shows the region of the landscape explored by Figures 2-3.
@@
 Other support landscapes would have very different implications for the results they obtained.
-A small fraction of analyses may have supported Porifera-sister because they happened to sample a small island of support (Figure 4b) in a broader landscape that overall has patterns of support consistent with Ctenophora-sister.
+A small fraction of analyses may have supported Porifera-sister because they happened to sample a small island of support (Figure 5b) in a broader landscape that overall has patterns of support consistent with Ctenophora-sister.
 Alternatively, the small island of support for Porifera-sister may indicate a transition from a plateau of support for Ctenophora sister to a hilly support landscape to the east.

### 2017-12-12 readme.rmd
typo

@@
 These are the six recoded CAT-GTR+G analyses with posterior predictive (PP) scores.
-The presented statistics for these six cells are posterior probability of Poifera-sister, Maxdiff (with lower scores indicating better convergence), and five PP scores (where lower absolute value indicates better model adequacy).
+The presented statistics for these six cells are posterior probability of Porifera-sister, Maxdiff (with lower scores indicating better convergence), and five PP scores (where lower absolute value indicates better model adequacy).
 Only one analysis provides significant support for Porifera-sister.
@@
 By their own criterion of model adequacy, we should favor the unresolved analyses.
+Within these recoded CAT-GTR+G analyses, there is not a trend toward increased support for Porifera-sister with increasing model adequacy.
 

### 2017-12-13 readme.rmd
adjusted figure legend

@@
 This left six cells that are the focus of their primary conclusions (Figures 2-3, right panel, top two rows, right 3 cells).
+A variety of statistics from Feuda *et al.* 2017 for these six cells are presented in Figure 4.
 
@@
 The presented statistics for these six cells are posterior probability of Porifera-sister, Maxdiff (with lower scores indicating better convergence), and five PP scores (where lower absolute value indicates better model adequacy).
-Only one analysis provides significant support for Porifera-sister.
-It isn't the most adequate by any of the PP scores, and showed the poorest convergence according to Maxdiff.
-Recodings are: None- K - KGB-6, D - Dayhoff-6, and S- SR-6.*
+Only one analysis provides significant support for Porifera-sister (shown in green).
+It is not the most adequate analysis by any of the PP scores, and showed the poorest convergence according to Maxdiff.
+Recodings are: K - KGB-6, D - Dayhoff-6, and S- SR-6.*
 
-A variety of statistics from Feuda *et al.* 2017 for these six cells are presented in Figure 4.
 The first thing to note is that only one of these six analyses provides significant support for Porifera-sister.

### 2018-01-08 readme.rmd
reworded

@@
 - Most analyses under the CAT-GTR+G model do not provide significant support for either Ctenophora-sister or Porifera-sister (Figure 3, right panel, 10 blue cells)
-- A subset of matrices with reduced species sampling and recoded data provide significant support for Porifera-sister (Figure 3, right panel, 4 green cells)
+- A subset of matrices with recoded data, most of which also have reduced species sampling, provide significant support for Porifera-sister (Figure 3, right panel, 4 green cells)
 

### 2018-04-10 readme.rmd
more details on ppa

@@
 
-The primary explanation Feuda *et al.* 2017 provide for accepting some results and rejecting others is their posterior predictive (PP) analyses of model adequacy.
-They calculated PP for only the Whelan-Opisthokont and Chang matrices (Figures 2-3, top two rows).
-Based on these scores they rejected the results of the GTR-G analyses (Figures 2-3, left pane) and analyses that had not been recoded.
-This left six cells that are the focus of their primary conclusions (Figures 2-3, right panel, top two rows, right 3 cells).
-A variety of statistics from Feuda *et al.* 2017 for these six cells are presented in Figure 4.
+The primary criteria Feuda *et al.* 2017 provide for accepting some results and rejecting others is their posterior predictive (PP) analyses of model adequacy ([Bollback 2002](https://doi.org/10.1093/oxfordjournals.molbev.a004175)).
+Adequacy is assessed in this context by assessing how similar data simulated from the posterior distribution under the model are to the actual data.
+This is used as a proxy for the the ability of the model to predict future observations.
+Each PP analysis results in a a score indicating how different the simulated data are from the actual data.
+PP analyses with scores closer to 0 are interpreted as being more adequate.
+
+Different PP statistics have been developed to consider the adequacy of the models to predict different aspects of the data.
+Feuda *et al.* 2017 calculated five different PP statistics with the program phylobayes-mpi:
+
+- ppa_div (phylobayes-mpi flag `-div`, output `diversity test`). Assesses the ability to predict the mean number of states at each site.
+
+- ppa_conv (phylobayes-mpi flag `-siteconvprob`, output `empirical convergence probability test`).
+Assesses the ability to predict the probability of convergence toward the same state at a given site in different taxa.
+
+- ppa_var (phylobayes-mpi flag `-sitecomp`, output `across-site compositional heterogeneity test`). Assesses the ability to predict compositional heterogeneity of states across sites.
+
+- ppa_max (phylobayes-mpi flag `-comp`, output `max heterogeneity across taxa`). Assesses the ability to predict compositional heterogeneity of states across taxa.
+
+- ppa_mean (phylobayes-mpi flag `-comp`, output `mean squared heterogeneity across taxa`). Assesses the ability to predict compositional heterogeneity of states across taxa.
+
+Feuda *et al.* 2017 calculated these PP statistics for only the Whelan-Opisthokont and Chang matrices (Figures 2-3, top two rows).
+Based on PP scores they rejected the results of the GTR-G analyses (Figures 2-3, left pane).
+This leaves the eight cells shown in Figure 4, presented here with a variety of statistics from Feuda *et al.* 2017.
+
+The left column is analyses that have not been recoded, and the other three columns present the analyses for the three types of recoding.
+Feuda *et al.* 2017 reject the analyses that have not been recoded, all of which provide significant support for Ctenophora-sister, based on PP scores.
+These PP scores do not, however, uniformly indicate improved adequacy of recoded analyses over analyses that have not been recoded.
+Specifically, the ppa-conv analyses suggest that the analyses that have not been recoded are more adequate (ie, have a lower score) than the recoded analyses.
+This is the case for both the Chang and Whelan matrix.
+The impact of recoding on PP analyses therefore depends on the  PP statistic under consideration.
 
@@
 	filter( (Dataset=="Whelan-Opis") | (Dataset=="Chang-Holo") ) %>%
-	filter( Recoding != "None" ) %>%
 	select (-c(Ctenophora_sister, Burnin, effsize, rel_diff)) %>%
@@
 			facet_wrap( ~ var, ncol = 2 ) +
-			geom_text(aes(label = val)) +
-			scale_fill_manual( values=c( poro_color,unresolved_color ) ) +
+			geom_text(aes(label = val), size=3) +
+			scale_fill_manual( values=c( cteno_color, poro_color,unresolved_color ) ) +
 			ggtitle("Variation between analyses") +
@@
 *Figure 4 - Detailed examination of the analysis subset (indicated in upper right pane) that is the focus of the conclusions by Feuda et al.
-These are the six recoded CAT-GTR+G analyses with posterior predictive (PP) scores.
+These include the six recoded CAT-GTR+G analyses with posterior predictive (PP) scores.
 The presented statistics for these six cells are posterior probability of Porifera-sister, Maxdiff (with lower scores indicating better convergence), and five PP scores (where lower absolute value indicates better model adequacy).
@@
 
-The first thing to note is that only one of these six analyses provides significant support for Porifera-sister.
+Though the major conclusion of Feuda *et al.* is that their "results strongly support sponges as the sister group of all other animals", only one of these analyses provides significant support for Porifera-sister.
 This is for the SR-6 recoded Whelan-Opisthokont matrix, shown in green in Figure 4.
-Given the similarity of these six analyses, the fact that only one supports Porifera-sister calls into serious question the robustness of support for Porifera-sister.
+The fact that only one of these analyses provides significant support for Porifera-sister calls into serious question the robustness of support for Porifera-sister.
 
@@
 The one analysis with significant support for Porifera-sister does not have the best (lowest absolute value) PP score for any of the five PP statistics considered (Figure 4).
-By their own criterion of model adequacy, we should favor the unresolved analyses.
-Within these recoded CAT-GTR+G analyses, there is not a trend toward increased support for Porifera-sister with increasing model adequacy.
+By their own criterion of model adequacy, we should favor either analyses that provide strong significant support for Ctenophora-sister (if ppa_conv is considered) or unresolved analyses (if any of the other four PP statistics are considered).
+There is not a trend toward increased support for Porifera-sister with increasing model adequacy.
 
-In fact, the only statistic that stands out for the significant analysis is the high Maxdiff score, which indicates it is the most poorly converged of these six focal analyses.
+The only statistic that stands out for the one analysis with PP scores that provides significant support for Porifera-sister is a high Maxdiff score (Figure 4).
+This score indicates it is the most poorly converged of the recoded analyses.
 

### 2018-06-18 readme.rmd
updated ppa-var wording
changes

@@
 This is the case for both the Chang and Whelan matrix.
+The ppa-var analyses indicate that that analyses that have not been recoded are more adequate than analyses that have been recoded with Dayhoff-6  and SR-6 recoding.
 The impact of recoding on PP analyses therefore depends on the  PP statistic under consideration.
@@
 *Figure 4 - Detailed examination of the analysis subset (indicated in upper right pane) that is the focus of the conclusions by Feuda et al.
-These include the six recoded CAT-GTR+G analyses with posterior predictive (PP) scores.
-The presented statistics for these six cells are posterior probability of Porifera-sister, Maxdiff (with lower scores indicating better convergence), and five PP scores (where lower absolute value indicates better model adequacy).
+These are the eight CAT-GTR+G analyses with posterior predictive (PP) scores.
+The presented statistics for these cells are posterior probability of Porifera-sister, Maxdiff (with lower scores indicating better convergence), and five PP scores (where lower absolute value indicates better model adequacy).
 Only one analysis provides significant support for Porifera-sister (shown in green).

### 2018-06-19 manuscript.rmd
added manuscript draft

---
output:
  github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	message = FALSE,
	warning = FALSE,
	echo=FALSE,
	dpi=300,
	cache=TRUE
	)
```

```{r preliminaries}
library( tidyverse )
library( gridExtra )
library( ape )
library( hutan )
library( magrittr )
library( doParallel )
library( png )

# Set system computational parameters
cores = detectCores() - 1
if ( cores < 1 ) {
	cores = 1
}

# Register parallel workers for %dopar%
registerDoParallel( cores )

# set constants
burnin = 200 # burning for reading tree support

breaks = c( 20, 6 ) # breakpoints on plots

path_analyses = "recoding/alternative/"

all_legends = FALSE # set to true to validate color scheme

```

```{r read_ppred}

ppred_files = list.files( path=path_analyses, pattern = "\\.ppred$" )
ppred_files = paste( path_analyses, ppred_files, sep="" )

# Static file structure, but check everything is in the right place
expected_test_name_lines = c( 1, 7, 13, 21, 27 )
expected_test_names = c(
	"diversity test",
	"empirical convergence probability test",
	"across-site compositional heterogeneity test",
	"max heterogeneity across taxa",
	"mean squared heterogeneity across taxa"
)

expected_zscore_lines = expected_test_name_lines + 3

ppred =
	lapply(
		ppred_files,
		function( ppred_file ){
			row = str_match( ppred_file, "WhelanD20_AA_mod_transform-([0-9a-zA-Z]+)_states-([0-9]+).chain.ppred" )  # "\\((.*?) :: (0\\.[0-9
			dim(row) = NULL
			lines = read_lines( ppred_file )
			if ( ! all( lines[ expected_test_name_lines ] == expected_test_names ) ){
				stop("ppred file not of expected structure")
			}
			z_scores =
				lines[ expected_zscore_lines ] %>%
				str_replace( "z-score : ", "") %>%
				as.numeric()

			D = data.frame(
				file = row[ 1 ],
				transform = row[ 2 ],
				states = as.numeric( row[ 3 ] ),

				# The naming of the ppred scores isn't consistent between the phylobayes
				# manual, the phylobayes ppred output, and Feuda et al. Below are notes
				# used to reconcile these.

				# name in ppred file: diversity test
				# phylobaesmpi flag: -div, "performs a posterior predictive diversity test: the test statistic is the mean diversity per site"
				# Feuda et al name: ppa-div, "site-specific amino acid diversity... measuring whether models can adequately estimate... the mean number of distinct amino acids observed at each site"
				ppa_div = z_scores[ 1 ],

				# name in ppred file: empirical convergence probability test
				# phylobaesmpi flag: -siteconvprob
				# Feuda et al name: ppa-conv
				ppa_conv = z_scores[ 2 ],

				# name in ppred file: across-site compositional heterogeneity test
				# phylobaesmpi flag: -sitecomp
				# Feuda et al name: PPA-VAR
				ppa_var = z_scores[ 3 ],

				# name in ppred file: max heterogeneity across taxa
				# phylobaesmpi flag:
				# Feuda et al name: PPA-MAX
				ppa_max = z_scores[ 4 ],

				# name in ppred file: mean squared heterogeneity across taxa
				# phylobaesmpi flag:
				# Feuda et al name: PPA-MEAN
				ppa_mean = z_scores[ 5 ]
			)

			return( D )
		}
	) %>%
	bind_rows()

D =
	ppred %>%
	gather( ppred, score, ppa_div:ppa_mean)

#D %>%
#	ggplot( aes(x=states, y=score, group=transform, col=transform) ) +
#		geom_line()+
#		geom_point()+
#		scale_x_reverse() +
#		facet_grid( . ~ ppred )

D$transform = paste(rep("random-", length(D$transform)), D$transform, sep="")
D$transform[D$transform == "random-none"] = "SR6-reproduced"

```

```{r read_ppred_published}

original = 
	read.table("./tidy_table_3.tsv", sep="\t", header=T) %>%
	filter( Model == "CAT-GTR+G" ) %>%
	filter( Recoding == "SR-6" | Recoding == "None" ) %>%
	filter( Dataset == "WhelanD20-Opistho" ) %>%
	na.omit() %>% 
	mutate( states = ifelse(Recoding=="None", 20, 6) ) %>%
	gather( ppred, score, PPA_DIV:PPA_MEAN) %>%
	mutate( ppred = str_to_lower(ppred) ) %>% 
	select( states:score ) %>%
	mutate( transform = "SR6-published" )

D = bind_rows( D, original )

D$transform = as.factor(D$transform)

#use more descriptive PP names
D$ppred[D$ppred=="ppa_div"] = "PP diversity"
D$ppred[D$ppred=="ppa_conv"] = "PP convergence"
D$ppred[D$ppred=="ppa_var"] = "PP site hetero"
D$ppred[D$ppred=="ppa_max"] = "PP taxon hetero max"
D$ppred[D$ppred=="ppa_mean"] = "PP taxon hetero mean"

```

```{r read_trees}

# Read in trees for each run, calculate cteno-sis and porifera-sis support

fungi = c(
	"Rhizopus_oryzae", 
	"Mortierella_verticillata", 
	"Allomyces_macrogynus", 
	"Spizellomyces_punctatus"  )

porifera = c(
	"Aphrocallistes_vastus",
	"Rossella_fibulata",
	"Sympagella_nux",
	"Hyalonema_populiferum",
	"Ircinia_fasciculata",
	"Chondrilla_nucula",
	"Latrunculia_apicalis",
	"Kirkpatrickia_variolosa",
	"Crella_elegans",
	"Tethya_wilhelma",
	"Pseudospongosorites_suberitoides",
	"Ephydatia_muelleri",
	"Spongilla_alba",
	"Amphimedon_queenslandica",
	"Petrosia_ficiformis",
	"Sycon_ciliatum",
	"Sycon_coactum",
	"Oscarella_carmela",
	"Corticium_candelabrum"
)

ctenophora = c(
	"Euplokamis_dunlapae",
	"Pleurobrachia_bachei",
	"Pleurobrachia_atlantica",
	"Dryodora_glandiformis",
	"Bolinopsis_infundibulum",
	"Mertensiidae_sp",
	"Mnemiopsis_leidyi",
	"Beroe_abyssicola",
	"Vallicula_sp",
	"Coeloplana_astericola"
)

animals_non_poro_cteno = c(
	"Trichoplax_adhaerens", 
	"Petromyzon_marinus", 
	"Danio_rerio", 
	"Homo_sapiens", 
	"Hemithiris_psittacea", 
	"Tubulanus_polymorphus", 
	"Capitella_teleta", 
	"Lottia_gigantea", 
	"Priapulus_caudatus", 
	"Lithobius_forficatus", 
	"Drosophila_melanogaster", 
	"Daphnia_pulex", 
	"Strongylocentrotus_purpuratus", 
	"Saccoglossus_kowalevskii", 
	"Platygyra_carnosa", 
	"Acropora_digitifera", 
	"Aiptasia_pallida", 
	"Hormathia_digitata", 
	"Bolocera_tuediae", 
	"Nematostella_vectensis", 
	"Eunicella_verrucosa", 
	"Aurelia_aurita", 
	"Periphylla_periphylla", 
	"Physalia_physalis", 
	"Nanomia_bijuga", 
	"Agalma_elegans", 
	"Craseoa_lathetica", 
	"Abylopsis_tetragona", 
	"Hydra_viridissima", 
	"Hydra_vulgaris", 
	"Hydra_oligactis"
)

get_clade_frequency = function( trees, clade ){
	x = lapply(trees, function(tree){
		is_monophyletic(tree, clade)
	}) %>% 
		unlist %>% 
		sum

	return( x / length(trees) )

}

#test_name = "WhelanD20_AA_mod_transform-01_states-14.chain.treelist"
#trees = read.tree(test_name)
#burnin = 200
#trees = trees[(burnin+1):length(trees)]

#get_clade_frequency(trees, clade_ctenosis)
#get_clade_frequency(trees, clade_porisis)

clade_ctenosis = c( animals_non_poro_cteno, porifera )
clade_porisis  = c( animals_non_poro_cteno,  ctenophora )

tree_files  = list.files( path=path_analyses, pattern = "\\.treelist$" )
tree_files = paste( path_analyses, tree_files, sep="" )

get_tree_support = function( tree_file ){

	row = str_match( tree_file, "WhelanD20_AA_mod_transform-([0-9a-zA-Z]+)_states-([0-9]+).chain.treelist" )  # "\\((.*?) :: (0\\.[0-9
	dim(row) = NULL

	trees = read.tree(tree_file)
	burnin = 200
	trees = trees[(burnin+1):length(trees)]

	ctenosis = get_clade_frequency(trees, clade_ctenosis)
	porisis = get_clade_frequency(trees, clade_porisis)
	D = data.frame(
		file = row[ 1 ],
		transform = row[ 2 ],
		states = as.numeric( row[ 3 ] ),
		ctenosis = ctenosis,
		porisis = porisis
	)

	return( D )
}

tree_support = foreach( tree_file=tree_files ) %dopar%
	get_tree_support( tree_file )

tree_support = tree_support %>% bind_rows()

# Rename transforms

tree_support$transform = paste(rep("random-", length(tree_support$transform)), tree_support$transform, sep="")
tree_support$transform[tree_support$transform == "random-none"] = "SR6-reproduced"

# Add published results, as hard coded from Feuda et al.

pub06 = data.frame(
	file = NA,
	transform = "SR6-published",
	states = 6,
	ctenosis = 0.0,
	porisis = 0.98,
	stringsAsFactors=F )

tree_support = tree_support %>% bind_rows( pub06 )

pub20 = data.frame(
	file = NA,
	transform = "SR6-published",
	states = 20,
	ctenosis = 1.0,
	porisis = 0.00,
	stringsAsFactors=F )

tree_support = tree_support %>% bind_rows(  pub20 )

tree_support$transform = factor( tree_support$transform, levels=levels(D$transform) )
```

```{r plots_pp}
plot_ppa = function( ppred_focal, D, show_legend=FALSE ){
	g_plot = 
		D %>%
			filter( ppred==ppred_focal ) %>% 
			ggplot( aes(x=states, y=score, group=transform, col=transform) ) +
			geom_line()+
			geom_point()+
			scale_x_reverse(breaks=breaks) +
			ggtitle(ppred_focal) +
			expand_limits(y=c(0, NA)) +
			theme_classic()

	if( show_legend ){
		return( g_plot )
	}
	else{
		return( g_plot + theme(legend.position="none") )
	}

}

# From https://stackoverflow.com/questions/11883844/inserting-a-table-under-the-legend-in-a-ggplot2-histogram
g_legend = function(a.gplot){ 
	tmp = ggplot_gtable(ggplot_build(a.gplot) )
	leg = which(sapply(tmp$grobs, function(x) x$name) == "guide-box") 
	legend = tmp$grobs[[leg]] 
	return( legend )}

# g_ppa = sapply( unique(D$ppred), function(ppred){ plot_ppa( ppred, D ) } )

ppreds = unique(D$ppred)

#grid.arrange(
#	plot_ppa( ppreds[1], D, all_legends),
#	plot_ppa( ppreds[2], D, all_legends),
#	plot_ppa( ppreds[3], D, all_legends),
#	plot_ppa( ppreds[4], D, all_legends),
#	plot_ppa( ppreds[5], D, all_legends),
#	g_legend( plot_ppa( ppreds[5], D, TRUE ) ),
#	nrow = 2,
#	ncol = 3
#)

```

```{r figure_1, fig.width=6.85 }

states_to_consider = c( 20, 6)

D %<>% filter( states %in% states_to_consider)
tree_support %<>% filter( states %in% states_to_consider)

plot_tree_support = 
	tree_support %>% 
		ggplot( aes(x=states, y=porisis, group=transform, col=transform) ) +
		geom_line()+
		geom_point()+
		scale_x_reverse(breaks=breaks) +
		theme_classic() + 
		theme(legend.position="none") +
		ggtitle("Porifera Sister") +
		ylab("Posterior probability")

lay <- rbind(c(NA,NA,1),
						 c(2,3,4),
						 c(5,6,7))

g = arrangeGrob(
	g_legend( plot_ppa( ppreds[5], D, TRUE ) ),
	plot_tree_support,
	plot_ppa( ppreds[3], D, all_legends),
	plot_ppa( ppreds[2], D, all_legends),
	plot_ppa( ppreds[1], D, all_legends),
	plot_ppa( ppreds[4], D, all_legends),
	plot_ppa( ppreds[5], D, all_legends),
	layout_matrix = lay
)

# grid.arrange( g )

ggsave( file="figures/Figure1_manuscript_raw.pdf", device="pdf", width = 6.5, height = 4, dpi = 300, g )
```

# Response to Feuda *et al.*

Casey W. Dunn

## Summary

In the past decade, some analyses have placed Ctenophora (comb jellies) and others Porifera (sponges) as the sister group to all other animals (Figure 1A). Feuda *et al.* [1] present new phylogenetic analyses that they claim provide conclusive support for the Porifera-sister hypothesis. Their analyses consider datasets from studies by Chang *et al.* [2] and Whelan *et al.* [3], both of which found support for Ctenophora-sister. Feuda *et al.* were concerned that these Ctenophora-sister results were artefacts of lineage-specific differences in amino acid frequencies. In an attempt to reduce these differences, they recoded the full set of twenty amino acids into six groups of amino acids. These groups have more frequent evolutionary changes within them than between them, based on empirical observations in large protein datasets [4]. The intent is to discard many lineage-specific changes, which are expected to fall within these groups. They report that posterior predictive (PP) analyses [5] indicate recoded analyses have better model adequacy than amino acid analyses, and "Porifera-sister was favored under all recoding strategies". Here I focus on two aspects of Feuda *et al.* First, I point out that many of their recoded analyses are actually unresolved, and that the analyses with the best PP scores do not provide significant support for Porifera-sister. Second, I present new analyses that show that the impact of recoding is largely due to discarding information, not accommodating variation in amino acid composition across lineages. These findings indicate that it is premature to accept Porifera-sister and reject Ctenophora-sister. The new analyses show that recoding can be a problematic method for addressing compositional variation in phylogenetic analyses.

## Main Text

Feuda *et al.* examine support for Ctenophora-sister and Porifera-sister under all combinations of two models of molecular evolution, four datasets, and four coding schemes. This provides 32 analyses that they report in their Table 3 and that I present graphically here as Supplemental Figure 1. There is striking variation in support for Ctenophora-sister and Porifera-sister across these analyses (Supplemental Figure 1). Feuda *et al.* accept the results of some analyses and reject others based on posterior predictive (PP) analyses of model adequacy, which assess how well a model explains variation in the data (Bollback 2002). They considered five different PP statistics that capture different types of variation in the data, such as the heterogeneity of states across taxa (PP taxon hetero in figures here) and sites (PP site hetero). From this they conclude that their "results strongly support sponges as the sister group of all other animals".

This conclusion does not follow from their own presented results. Only a single analysis with PP scores provides what could be considered strong support (>95% posterior probability) for Porifera-sister. Of the 32 analyses, PP scores were calculated for 16 (those for the full Whelan and Chan matrices). Based on PP scores, Feuda *et al.* reject eight of these that were conducted under the GTR+G model (which all have strong support for Ctenophora-sister). This leaves eight CAT-GTR+G analyses (Supplemental Figure 1). Two of these eight are analyses of the original 20-state amino acid data, both of which provide strong support for Ctenophora-sister. Of the six recoded analyses, five are unresolved. Only a single analysis for which PP scores are available provides significant support for Porifera-sister, the CAT-GTR+G analysis of the SR-6 [4] recoded Whelan [3] matrix. Furthermore, this analysis does not have the best score according to any of the five PP statistics they considered (Supplemental Figure 1). The only statistic that stands out for this one analysis is that it has the highest maxdiff (Supplemental Figure 1), indicating that it did not converge as well as other analyses.

Though their study does not provide strong support for Porifera-sister, the sensitivity of their results to recoding provides an opportunity to better understand and evaluate the impact of recoding more generally. This is important given the growing interest in recoding [1]. Feuda *et al.* hoped recoding would reduce potential artefacts due to differences across species in amino acid frequencies. They interpreted the fact that their analyses are sensitive to recoding as evidence that such an artefact exists and that they successfully addressed it by recoding. An alternative hypothesis is that recoding impacts phylogenetic analyses simply because it discards so much information. These two hypotheses can be tested by applying new recoding schemes that also reduce twenty states down to six, but are based on random grouping rather than empirical frequencies of amino acid exchange. Empirical and random recodings both discard the same amount of information, but only empirical recoding could reduce the impact of amino-acid frequency as intended. Different results between empirical and random recoding would be consistent with the hypothesis that the empirical approach works as intended to accommodate differences in amino acid frequencies across species. Similar results would suggest that the impact of recoding on phylogenetic results is due to discarding data. Here I focus on the single original analysis with a PP score that supports Porifera-sister, the CAT-GTR+G analysis of the SR-6 recoded Whelan data. I created four new random recoding schemes by shuffling the amino acids in the SR-6 scheme (analysis code is available in a git repository at https://github.com/caseywdunn/feuda_2017). When I applied each of these randomized codes to the Whelan matrix and analyzed them under the CAT-GTR+G model with phylobayes-mpi [6], I observed similar results as for the empirical SR-6 recoding. Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Figure 1B).

These analyses indicate that the major impact of recoding on phylogenetic analyses is data reduction, not reduction of compositional variation across species. This suggests that recoding may not be an effective tool for accommodating among-species differences in amino acid frequencies more generally. Compositional heterogeneity would be better addressed with models of molecular evolution that explicitly describe such differences [7,8], if progress can be made on the considerable computational challenges of such complex models.

### Author Contributions
CWD performed all analyses and wrote the manuscript.

## References

1.	Feuda, R., Dohrmann, M., Pett, W., Philippe, H., Rota-Stabelli, O., Lartillot, N., Wörheide, G., and Pisani, D. (2017). Improved Modeling of Compositional Heterogeneity Supports Sponges as Sister to All Other Animals. Curr. Biol. 27, 3864–3870.e4.
2.	Chang, E.S., Neuhof, M., Rubinstein, N.D., Diamant, A., Philippe, H., Huchon, D., and Cartwright, P. (2015). Genomic insights into the evolutionary origin of Myxozoa within Cnidaria. Proc. Natl. Acad. Sci. U. S. A. 112, 14912–14917.
3.	Whelan, N.V., Kocot, K.M., Moroz, L.L., and Halanych, K.M. (2015). Error, signal, and the placement of Ctenophora sister to all other animals. Proc. Natl. Acad. Sci. U. S. A. 112, 5773–5778.
4.	Susko, E., and Roger, A.J. (2007). On reduced amino acid alphabets for phylogenetic inference. Mol. Biol. Evol. 24, 2139–2150.
5.	Bollback, J.P. (2002). Bayesian model adequacy and choice in phylogenetics. Mol. Biol. Evol. 19, 1171–1180.
6.	Lartillot, N., Rodrigue, N., Stubbs, D., and Richer, J. (2013). PhyloBayes MPI: phylogenetic reconstruction with infinite mixtures of profiles in a parallel environment. Syst. Biol. 62, 611–615.
7.	Blanquart, S., and Lartillot, N. (2008). A site- and time-heterogeneous model of amino acid replacement. Mol. Biol. Evol. 25, 842–858.
8.	Foster, P.G. (2004). Modeling compositional heterogeneity. Syst. Biol. 53, 485–495.

### Figure

![](figures/Figure1_manuscript.png)
**Figure 1.** (A) The two alternative hypotheses for deep animal relationships considered here. Relationships that are not part of these hypotheses are shown as unresolved polytomies. (B) Each of the six plots presents one statistic, which include Posterior probability of Porifera-sister and the five Posterior Predictive (PP) statistics considered by Feuda *et al.* Within each plot, there are six lines for six different analyses. These six analyses are the published SR-6 analyses presented by Feuda *et al.* (SR6-published), reproduced analyses obtained by applying the same methods to the same data (SR6-reproduced), and four analyses based on randomized recoding matrices obtained by shuffling the SR-6 coding scheme (random-00 – random-03). Each analysis includes results for 20 states (the raw amino acid data, shown by the left point) and for 6 states (the 6-state recoded data, shown by the right point). For each statistic, the results obtained with the random recoding are similar to those of the SR6 recoding. This indicates that the impact of recoding is dominated by discarding data when collapsing from 20 states to 6 states, not accommodating compositional heterogeneity across lineages.

\pagebreak

## Supplemental Information

### Supplemental methods

All code used for the analyses presented here is available in a git repository at https://github.com/caseywdunn/feuda_2017. The version of the code executed to present the results here has tag `cb_submission`. The randomized recoding analyses are in the `recoding/alternative` subdirectory of the git repository.

The original SR-6 recoding scheme is "`APST CW DEGN FHY ILMV KQR`" [4], where spaces separate amino acids that are placed into the same group.

The `alt_recode.py` script was used to generate the randomized recoding schemes and apply the recodings to the data. To create the randomized recoding schemes, the amino acids in the SR-6 recoding were randomly reshuffled. This creates new recodings that have the same sized groups as SR-6. The new recodings were, from `random-00` to `random-03`:

    MPKE AY IDGQ TRC WNLF SVH
    EIFT WL QVPG NKM SCHA RYD
    LCQS GK WPTI VRD YEFN MAH
    IWQV TY KDLM ESH APCF GRN

To apply these to the data, each amino acid was replaced with the first amino acid in the group. When applying `random-00`, for example, each instance of `R` and `C` would be replaced by a `T`.

The 20 state matrices are the same across all analyses since they are not recoded. Since all the 20 state matrices are all the same, variation between 20-state results (as in the left side of each pane of Figure 1B) give insight into the technical variance of the inference process.

Each new matrix was analyzed with phylobayes-mpi version 1.8. Analyses were run for 1000 generations, and a 200 generation burnin applied. The resulting tree files and PP scores were parsed for display with the code in https://github.com/caseywdunn/feuda_2017/manuscript.rmd .

The statistics presented in Supplemental Figure 1 were parsed from the Feuda *et al.* manuscript into the file `tidy_table_3.tsv` and rendered for display in R.

### Supplemental figure

![](figures/Supplemental_Figure_1_manuscript.png)

**Supplemental Figure 1.** The subset of eight CAT-GTR+G analyses with posterior predictive (PP) scores that is the focus of Feuda *et al.*'s primary conclusions. These are a subset of the 32 analyses presented in their Table 3 and graphically here in the upper right pane. The eight analyses are for two datasets (Chang and Whelan) and four coding schemes. The coding schemes are the original 20 state amino acid data, and three different six state recodings that group amino acids based on different criteria: KGB6, D6, and SR6. Cells are color coded according to whether support for Ctenophora-sister (red) or Porifera-sister (green) is greater than 95% posterior probability, or unresolved (the posterior probabilities for neither hypothesis are greater than 95%). Only one of these analyses, the SR6 coding of the Whelan matrix, has >95% support for Porifera-sister. The 20-state and 6-state points on the plots in Figure 1 of the main text correspond to the 20 and SR6 Whelan cells shown here. The presented statistics for these cells are posterior probability of Porifera-sister, Maxdiff (with lower scores indicating better convergence of runs), and five PP statistics (where lower absolute value indicates better model adequacy). The only one of these eight analyses that provides significant support for Porifera-sister is not the most adequate analysis by any of the PP scores, and showed the poorest convergence according to Maxdiff.

### 2018-06-19 recoding/alternative/alt_recode.py

@@ -1,0 +1,199 @@
+import sys
+import re
+from string import maketrans
+import random
+import collections
+
+# Tools to recode phylip data matrices, ie to substitute every occurance of a
+# particular state with another state.
+#
+# These tools support a reduced representation recoding scheme, as well as
+# transforms that modify the recoding scheme.
+#
+# Here is how I refer to these processes:
+# - Recoding is the application of a reduced state recoding map that results
+#   in a smaller number of character states.
+#   A Recoding is specified by a space delimeted string called the code, where
+#   all the states to be mapped to the same state are in an uniterrupted block.
+#   The first state in each block is used as the new state for all states in
+#   the block. This code is converted to a map that defines the recoding.
+#   This map is a tuple that contains two strings of equal length. The first
+#   has all the original states, and the second has the reduced states that
+#   each of those states map to (as defined by conserved position).
+# - Transforming is the application of a map that changes some states to other
+#   states. It results in the same number of states after as before the
+#   transform. It is used to scramble states. A transform is defined by a
+#   tuple that contains two strings of equal length. The first has all the
+#   original states, and the second has the states that each of those states
+#   map to (as defined by conserved position). These tools apply the transform
+#	to the recoding code, rather than to the data. The transform is useful to
+#	build up null expectations under random recodings.
+
+
+def sr_codes():
+	# The SR codings are from the left side of Table 1 of
+	# https://academic.oup.com/mbe/article-lookup/doi/10.1093/molbev/msm144
+	#
+	# Put an empty string in position 0 so that the index of the tuple
+	# corresponds to the number of recoded states
+	return (
+	"",
+	"ADEGKNPQRSTCFHILMVWY",
+	"ADEGKNPQRST CFHILMVWY",
+	"ADEGNPST CHKQRW FILMVY",
+	"AGNPST CHWY DEKQR FILMV",
+	"AGPST CFWY DEN HKQR ILMV",
+	"APST CW DEGN FHY ILMV KQR",
+	"AGST CW DEN FY HP ILMV KQR",
+	"AST CG DEN FY HP ILV KQR MW",
[…]

### 2018-06-19 recoding/alternative/phylobayesmpi_batch.sh

@@ -1,0 +1,20 @@
+#!/bin/bash
+#SBATCH --job-name=jtt_postpred
+
+#SBATCH --partition=general
+#SBATCH --constraint=avx2
+#SBATCH --ntasks=100
+
+#SBATCH --mem-per-cpu=6G
+#SBATCH --time=7-00:00:00
+
+
+module load PhyloBayes-MPI
+
+for filename in ./*_mod_*.phy; do
+  filename=$(basename "$filename")
+	base="${filename%.*}"
+	echo "processing $filename"
+	srun --mpi=pmi2 pb_mpi  -x 1 5000 -cat -gtr -dgam 4 -d $filename $base.chain
+	srun --mpi=pmi2 readpb_mpi -allppred -x 200 5 $base.chain > $base.ppred.out
+done

### 2018-06-20 manuscript.rmd
added a sentence
added additional sup fig

@@
 library( png )
+library( scales )
 
@@
 
+posterior_prob_thresh = 0.95
+
 breaks = c( 20, 6 ) # breakpoints on plots
@@
 all_legends = FALSE # set to true to validate color scheme
+
+```
+
+```{r load_table3_data}
+
+D = read.table( "tidy_table_3.tsv", header=TRUE, stringsAsFactors=FALSE  )
+
+D$Model = factor( D$Model, c("GTR+G", "CAT-GTR+G" ) )
+
+D$Dataset = sub( "D20", "", D$Dataset )
+D$Dataset = sub( "Chang", "Chang-Holo", D$Dataset )
+D$Dataset = sub( "Opistho", "Opis", D$Dataset )
+D$Dataset = sub( "Choano", "Cho", D$Dataset )
+D$Dataset = factor( D$Dataset, c( "Whelan-Cho", "Whelan-Holo", "Whelan-Opis", "Chang-Holo" ) )
+
+D$Recoding[ D$Recoding == "None" ] = "20"
+D$Recoding[ D$Recoding == "KGB-6" ] = "KGB6"
+D$Recoding[ D$Recoding == "Dayhoff-6" ] = "D6"
+D$Recoding[ D$Recoding == "SR-6" ] = "SR6"
+D$Recoding = factor( D$Recoding, c("20", "KGB6", "D6", "SR6" ) )
+
+D$Result = "Unresolved"
+D$Result[ D$Porifera_sister >= posterior_prob_thresh ] = "Porifera-sister"
+D$Result[ D$Ctenophora_sister >= posterior_prob_thresh ] = "Ctenophora-sister"
+D$Result = factor( D$Result )
+
+D$Convergence = "bad"
+D$Convergence[ D$Maxdiff <= 1 ] = "possible"
+D$Convergence[ D$Maxdiff <= 0.3 ] = "acceptable"
+D$Convergence[ D$Maxdiff <= 0.1 ] = "good"
+D$Convergence = factor( D$Convergence, c( "good", "acceptable", "possible", "bad" ) )
+
+D$Maxdiff = round( D$Maxdiff, 2 )
+
+cteno_color = "#F8766D"
+poro_color = "#00BA38"
+unresolved_color = "#6F94CD"
+
+```
+
+```{r result_support}
+
+gg_support = 
+D %>%
+	ggplot( aes( Recoding, Dataset ) ) +
+		geom_tile(aes(fill = Porifera_sister)) +
+		facet_wrap(~ Model) +
+		scale_fill_gradient(low = cteno_color, high = poro_color, limits=c(0, 1)) +
+		geom_text(aes(label = Porifera_sister)) +
+		ggtitle("Porifera-sister Support") +
+		theme_classic()
+
+ggsave( 
+	file="figures/Supplemental_Figure_1_manuscript_raw.pdf", 
+	device="pdf", 
+	width = 6.85, 
+	height = 4, 
+	dpi = 300, 
+	gg_support )
+
+```
+
+```{r result_sub}
+
+D_sub =
+	D %>%
+	rename( Porifera_sister_posterior = Porifera_sister ) %>%
+	filter(Model=="CAT-GTR+G") %>%
+	filter( (Dataset=="Whelan-Opis") | (Dataset=="Chang-Holo") ) %>%
+	select (-c(Ctenophora_sister, Burnin, effsize, rel_diff)) %>%
+	gather( var, val, Porifera_sister_posterior:PPA_MEAN ) %>%
+	mutate( 
+		var = 
+			factor( 
+				var, 
+				levels = c(
+					"Porifera_sister_posterior", 
+					"Cycles", 
+					"Maxdiff", 
+					"PPA_DIV", 
+					"PPA_MAX", 
+					"PPA_CONV", 
+					"PPA_VAR", 
+					"PPA_MEAN")
+				)
+		)
+
+gg_sub =
+	D_sub %>%
+		ggplot( aes( Recoding, Dataset ) ) +
+			geom_tile( aes( fill = Result ) ) +
+			facet_wrap( ~ var, ncol = 2 ) +
+			geom_text(aes(label = val), size=3) +
+			scale_fill_manual( values=c( cteno_color, poro_color,unresolved_color ) ) +
+			ggtitle("Variation between analyses") +
+			theme_classic()
+
+ggsave( 
+	"figures/Supplemental_Figure_2_manuscript_raw.pdf", 
+	gg_sub, 
+	device=pdf, 
+	width=6.5, 
+	height=6 
+	)
 
@@
 		function( ppred_file ){
-			row = str_match( ppred_file, "WhelanD20_AA_mod_transform-([0-9a-zA-Z]+)_states-([0-9]+).chain.ppred" )  # "\\((.*?) :: (0\\.[0-9
+			row = 
+				str_match( 
+					ppred_file, 
+					"WhelanD20_AA_mod_transform-([0-9a-zA-Z]+)_states-([0-9]+).chain.ppred" 
+					)  # "\\((.*?) :: (0\\.[0-9
+
 			dim(row) = NULL
@@
 
-			D = data.frame(
+			Dpp = data.frame(
 				file = row[ 1 ],
@@
 				# name in ppred file: diversity test
-				# phylobaesmpi flag: -div, "performs a posterior predictive diversity test: the test statistic is the mean diversity per site"
-				# Feuda et al name: ppa-div, "site-specific amino acid diversity... measuring whether models can adequately estimate... the mean number of distinct amino acids observed at each site"
+				# phylobaesmpi flag: -div, "performs a posterior predictive diversity test: 
+				#   the test statistic is the mean diversity per site"
+				# Feuda et al name: ppa-div, "site-specific amino acid diversity... 
+				#  measuring whether models can adequately estimate... the mean number of 
+				#  distinct amino acids observed at each site"
 				ppa_div = z_scores[ 1 ],
@@
 
-			return( D )
+			return( Dpp )
 		}
@@
 
-D =
+Dpp =
 	ppred %>%
@@
 
-#D %>%
+#Dpp %>%
 #	ggplot( aes(x=states, y=score, group=transform, col=transform) ) +
@@
 
-D$transform = paste(rep("random-", length(D$transform)), D$transform, sep="")
-D$transform[D$transform == "random-none"] = "SR6-reproduced"
+Dpp$transform = paste(rep("random-", length(Dpp$transform)), Dpp$transform, sep="")
+Dpp$transform[Dpp$transform == "random-none"] = "SR6-reproduced"
 
@@
 
-D = bind_rows( D, original )
+Dpp = bind_rows( Dpp, original )
 
-D$transform = as.factor(D$transform)
+Dpp$transform = as.factor(Dpp$transform)
 
 #use more descriptive PP names
-D$ppred[D$ppred=="ppa_div"] = "PP diversity"
-D$ppred[D$ppred=="ppa_conv"] = "PP convergence"
-D$ppred[D$ppred=="ppa_var"] = "PP site hetero"
-D$ppred[D$ppred=="ppa_max"] = "PP taxon hetero max"
-D$ppred[D$ppred=="ppa_mean"] = "PP taxon hetero mean"
+Dpp$ppred[Dpp$ppred=="ppa_div"] = "PP diversity"
+Dpp$ppred[Dpp$ppred=="ppa_conv"] = "PP convergence"
+Dpp$ppred[Dpp$ppred=="ppa_var"] = "PP site hetero"
+Dpp$ppred[Dpp$ppred=="ppa_max"] = "PP taxon hetero max"
+Dpp$ppred[Dpp$ppred=="ppa_mean"] = "PP taxon hetero mean"
 
@@
 	porisis = get_clade_frequency(trees, clade_porisis)
-	D = data.frame(
+	Dts = data.frame(
 		file = row[ 1 ],
@@
 
-	return( D )
+	return( Dts )
 }
@@
 
-tree_support$transform = factor( tree_support$transform, levels=levels(D$transform) )
+tree_support$transform = factor( tree_support$transform, levels=levels(Dpp$transform) )
 ```
@@
 ```{r plots_pp}
-plot_ppa = function( ppred_focal, D, show_legend=FALSE ){
+plot_ppa = function( ppred_focal, Dpp, show_legend=FALSE ){
 	g_plot = 
-		D %>%
+		Dpp %>%
 			filter( ppred==ppred_focal ) %>% 
@@
 
-# g_ppa = sapply( unique(D$ppred), function(ppred){ plot_ppa( ppred, D ) } )
+# g_ppa = sapply( unique(Dpp$ppred), function(ppred){ plot_ppa( ppred, Dpp ) } )
 
-ppreds = unique(D$ppred)
+ppreds = unique(Dpp$ppred)
 
 #grid.arrange(
-#	plot_ppa( ppreds[1], D, all_legends),
-#	plot_ppa( ppreds[2], D, all_legends),
-#	plot_ppa( ppreds[3], D, all_legends),
-#	plot_ppa( ppreds[4], D, all_legends),
-#	plot_ppa( ppreds[5], D, all_legends),
-#	g_legend( plot_ppa( ppreds[5], D, TRUE ) ),
+#	plot_ppa( ppreds[1], Dpp, all_legends),
+#	plot_ppa( ppreds[2], Dpp, all_legends),
+#	plot_ppa( ppreds[3], Dpp, all_legends),
+#	plot_ppa( ppreds[4], Dpp, all_legends),
+#	plot_ppa( ppreds[5], Dpp, all_legends),
+#	g_legend( plot_ppa( ppreds[5], Dpp, TRUE ) ),
 #	nrow = 2,
@@
 
-D %<>% filter( states %in% states_to_consider)
+Dpp %<>% filter( states %in% states_to_consider)
 tree_support %<>% filter( states %in% states_to_consider)
@@
 g = arrangeGrob(
-	g_legend( plot_ppa( ppreds[5], D, TRUE ) ),
+	g_legend( plot_ppa( ppreds[5], Dpp, TRUE ) ),
 	plot_tree_support,
-	plot_ppa( ppreds[3], D, all_legends),
-	plot_ppa( ppreds[2], D, all_legends),
-	plot_ppa( ppreds[1], D, all_legends),
-	plot_ppa( ppreds[4], D, all_legends),
-	plot_ppa( ppreds[5], D, all_legends),
+	plot_ppa( ppreds[3], Dpp, all_legends),
+	plot_ppa( ppreds[2], Dpp, all_legends),
+	plot_ppa( ppreds[1], Dpp, all_legends),
+	plot_ppa( ppreds[4], Dpp, all_legends),
+	plot_ppa( ppreds[5], Dpp, all_legends),
 	layout_matrix = lay
@@
 
-# grid.arrange( g )
+# grid.arrange( g ) # To render the plot directly in the text
 
@@
 The intent is to discard many lineage-specific changes, which are expected to fall within these groups.
+Rather than model compositional heterogeneity, this approach discards heterogeneous information so that much simpler models with fewer states can be applied.
 They report that posterior predictive (PP) analyses [5] indicate recoded analyses have better model adequacy than amino acid analyses, and "Porifera-sister was favored under all recoding strategies".
-Here I focus on two aspects of Feuda *et al.* First, I point out that many of their recoded analyses are actually unresolved, and that the analyses with the best PP scores do not provide significant support for Porifera-sister.
+Here I focus on two aspects of Feuda *et al.* First, I point out that many of their recoded analyses are actually unresolved, and that the analyses with the best PP scores do not provide strong support for Porifera-sister.
 Second, I present new analyses that show that the impact of recoding is largely due to discarding information, not accommodating variation in amino acid composition across lineages.
@@
 This conclusion does not follow from their own presented results.
-Only a single analysis with PP scores provides what could be considered strong support (>95% posterior probability) for Porifera-sister.
+Only a single analysis with PP scores provides what could be considered strong support (&ge; `r posterior_prob_thresh` posterior probability) for Porifera-sister.
 Of the 32 analyses, PP scores were calculated for 16 (those for the full Whelan and Chan matrices).
 Based on PP scores, Feuda *et al.* reject eight of these that were conducted under the GTR+G model (which all have strong support for Ctenophora-sister).
-This leaves eight CAT-GTR+G analyses (Supplemental Figure 1).
+This leaves eight CAT-GTR+G analyses (Supplemental Figure 2).
 Two of these eight are analyses of the original 20-state amino acid data, both of which provide strong support for Ctenophora-sister.
 Of the six recoded analyses, five are unresolved.
-Only a single analysis for which PP scores are available provides significant support for Porifera-sister, the CAT-GTR+G analysis of the SR-6 [4] recoded Whelan [3] matrix.
-Furthermore, this analysis does not have the best score according to any of the five PP statistics they considered (Supplemental Figure 1).
-The only statistic that stands out for this one analysis is that it has the highest maxdiff (Supplemental Figure 1), indicating that it did not converge as well as other analyses.
+Only a single analysis for which PP scores are available provides strong support for Porifera-sister, the CAT-GTR+G analysis of the SR-6 [4] recoded Whelan [3] matrix.
+Furthermore, this analysis does not have the best score according to any of the five PP statistics they considered (Supplemental Figure 2).
+The only statistic that stands out for this one analysis is that it has the highest maxdiff (Supplemental Figure 2), indicating that it did not converge as well as other analyses.
+It should also be noted that none of the reanalyses of Chang matrix provide strong contradiction of that original study.
+Chang *et al.* [2] found strong support for Ctenophora-sister, and all of the reanalyses of this same matrix by Feuda *et al.* also have strong support for Ctenophora-sister or are unresolved.
 
@@
 
-The statistics presented in Supplemental Figure 1 were parsed from the Feuda *et al.* manuscript into the file `tidy_table_3.tsv` and rendered for display in R.
+The statistics presented in Supplemental Figures 1 and 2 were parsed from the Feuda *et al.* manuscript into the file `tidy_table_3.tsv` and rendered for display with the code on the source file for this document.
 
-### Supplemental figure
+### Supplemental figures
 
-![](figures/Supplemental_Figure_1_manuscript.png)
+```{r result_sig, fig.width=6.5, fig.height=3.1}
 
-**Supplemental Figure 1.** The subset of eight CAT-GTR+G analyses with posterior predictive (PP) scores that is the focus of Feuda *et al.*'s primary conclusions.
+gg_sig =
+	D %>%
+		ggplot( aes( Recoding, Dataset ) ) +
+			geom_tile( aes( fill = Result ) ) +
+			facet_wrap( ~ Model ) +
+			theme_classic()
+
+#ggsave( "figures/figure_3.pdf", gg_sig, device=pdf, width=6.5, height=3.1 )
+
+gg_sig
+
+```
+**Supplemental Figure 1.**  A grapchical representation of the posterior probabilities for the 32 analyses presented by Feuda *et al.* in their Table 3.
+Cells are color coded by whether posterior probability is &ge; `r posterior_prob_thresh` for Porifera-sister,  &ge; `r posterior_prob_thresh` for Ctenophora-sister, or neither (unresolved).
+Posterior predictive (PP) statistics were estimated for the 16 analyses in the top two rows of this figure.
+
+![](figures/Supplemental_Figure_2_manuscript.png)
+
+**Supplemental Figure 2.** The subset of eight CAT-GTR+G analyses with posterior predictive (PP) scores that is the focus of Feuda *et al.*'s primary conclusions.
 These are a subset of the 32 analyses presented in their Table 3 and graphically here in the upper right pane.
@@
 The coding schemes are the original 20 state amino acid data, and three different six state recodings that group amino acids based on different criteria: KGB6, D6, and SR6.
-Cells are color coded according to whether support for Ctenophora-sister (red) or Porifera-sister (green) is greater than 95% posterior probability, or unresolved (the posterior probabilities for neither hypothesis are greater than 95%).
-Only one of these analyses, the SR6 coding of the Whelan matrix, has >95% support for Porifera-sister.
+Cells are color coded as in Supplemental Figure 1.
+Only one of these analyses, the SR6 coding of the Whelan matrix, has &ge; `r posterior_prob_thresh` support for Porifera-sister.
 The 20-state and 6-state points on the plots in Figure 1 of the main text correspond to the 20 and SR6 Whelan cells shown here.
 The presented statistics for these cells are posterior probability of Porifera-sister, Maxdiff (with lower scores indicating better convergence of runs), and five PP statistics (where lower absolute value indicates better model adequacy).
-The only one of these eight analyses that provides significant support for Porifera-sister is not the most adequate analysis by any of the PP scores, and showed the poorest convergence according to Maxdiff.
+The only one of these eight analyses that provides strong support for Porifera-sister is not the most adequate analysis by any of the PP scores, and showed the poorest convergence according to Maxdiff.

### 2018-06-20 manuscript.rmd
word smithing

@@
 The intent is to discard many lineage-specific changes, which are expected to fall within these groups.
-Rather than model compositional heterogeneity, this approach discards heterogeneous information so that much simpler models with fewer states can be applied.
+Rather than model compositional heterogeneity, as their title suggests, this approach discards heterogeneous information so that much simpler models with fewer states can be applied.
 They report that posterior predictive (PP) analyses [5] indicate recoded analyses have better model adequacy than amino acid analyses, and "Porifera-sister was favored under all recoding strategies".
@@
 The only statistic that stands out for this one analysis is that it has the highest maxdiff (Supplemental Figure 2), indicating that it did not converge as well as other analyses.
-It should also be noted that none of the reanalyses of Chang matrix provide strong contradiction of that original study.
+It should also be noted that none of the reanalyses of Chang matrix strongly contradict that study.
 Chang *et al.* [2] found strong support for Ctenophora-sister, and all of the reanalyses of this same matrix by Feuda *et al.* also have strong support for Ctenophora-sister or are unresolved.
@@
 These two hypotheses can be tested by applying new recoding schemes that also reduce twenty states down to six, but are based on random grouping rather than empirical frequencies of amino acid exchange.
-Empirical and random recodings both discard the same amount of information, but only empirical recoding could reduce the impact of amino-acid frequency as intended.
-Different results between empirical and random recoding would be consistent with the hypothesis that the empirical approach works as intended to accommodate differences in amino acid frequencies across species.
-Similar results would suggest that the impact of recoding on phylogenetic results is due to discarding data.
+Empirical and random recodings both discard the same amount of information, but only empirical recoding reduce the impact of amino-acid frequency as intended.
+Different results between empirical and random recoding would be consistent with the hypothesis that the empirical approach works as intended to accommodate compositional heterogeneity.
+Similar results would suggest that the impact of recoding is due to discarding information.
 Here I focus on the single original analysis with a PP score that supports Porifera-sister, the CAT-GTR+G analysis of the SR-6 recoded Whelan data.
-I created four new random recoding schemes by shuffling the amino acids in the SR-6 scheme (analysis code is available in a git repository at https://github.com/caseywdunn/feuda_2017).
+I created four new random recoding schemes by shuffling the amino acids in the SR-6 scheme (see Supplemental Methods and analysis code at https://github.com/caseywdunn/feuda_2017).
 When I applied each of these randomized codes to the Whelan matrix and analyzed them under the CAT-GTR+G model with phylobayes-mpi [6], I observed similar results as for the empirical SR-6 recoding.
@@
 
-These analyses indicate that the major impact of recoding on phylogenetic analyses is data reduction, not reduction of compositional variation across species.
+These analyses indicate that the major impact of recoding on phylogenetic analyses is data reduction, not accomodation of compositional heterogeneity across species.
 This suggests that recoding may not be an effective tool for accommodating among-species differences in amino acid frequencies more generally.
@@
 The original SR-6 recoding scheme is "`APST CW DEGN FHY ILMV KQR`" [4], where spaces separate amino acids that are placed into the same group.
+This recoding is one member of a family of recodings, each with a different number of groups, based on clustering of the JTT matrix.
+The other recoding used by Feuda *et al.*, KGB-6 and D-6, are based on different matrices and methods [1].
 
@@
 Each new matrix was analyzed with phylobayes-mpi version 1.8.
-Analyses were run for 1000 generations, and a 200 generation burnin applied.
+Analyses were run for 1000 generations, and a `r burnin` generation burnin applied.
 The resulting tree files and PP scores were parsed for display with the code in https://github.com/caseywdunn/feuda_2017/manuscript.rmd .

### 2018-06-20 manuscript.rmd
formatted citations

@@
 ---
+bibliography: manuscript.bib
+csl: current-biology.csl
 output:
@@
 
+Feuda *et al.* tackled a difficult phylogenetic problem -- placing the root of the animal phylogeny [@King:2017ie].
 In the past decade, some analyses have placed Ctenophora (comb jellies) and others Porifera (sponges) as the sister group to all other animals (Figure 1A).
-Feuda *et al.* [1] present new phylogenetic analyses that they claim provide conclusive support for the Porifera-sister hypothesis.
-Their analyses consider datasets from studies by Chang *et al.* [2] and Whelan *et al.* [3], both of which found support for Ctenophora-sister.
+Feuda *et al.* [@Feuda:2017ew] present new phylogenetic analyses that they claim provide conclusive support for the Porifera-sister hypothesis.
+Their analyses consider datasets from studies by Chang *et al.* [@Chang:2015hl] and Whelan *et al.* [@Whelan:2015jj], both of which found support for Ctenophora-sister.
 Feuda *et al.* were concerned that these Ctenophora-sister results were artefacts of lineage-specific differences in amino acid frequencies.
 In an attempt to reduce these differences, they recoded the full set of twenty amino acids into six groups of amino acids.
-These groups have more frequent evolutionary changes within them than between them, based on empirical observations in large protein datasets [4].
+These groups have more frequent evolutionary changes within them than between them, based on empirical observations in large protein datasets [@Susko:2007ds].
 The intent is to discard many lineage-specific changes, which are expected to fall within these groups.
 Rather than model compositional heterogeneity, as their title suggests, this approach discards heterogeneous information so that much simpler models with fewer states can be applied.
-They report that posterior predictive (PP) analyses [5] indicate recoded analyses have better model adequacy than amino acid analyses, and "Porifera-sister was favored under all recoding strategies".
+They report that posterior predictive (PP) analyses [@Bollback:2002to] indicate recoded analyses have better model adequacy than amino acid analyses, and "Porifera-sister was favored under all recoding strategies".
 Here I focus on two aspects of Feuda *et al.* First, I point out that many of their recoded analyses are actually unresolved, and that the analyses with the best PP scores do not provide strong support for Porifera-sister.
@@
 There is striking variation in support for Ctenophora-sister and Porifera-sister across these analyses (Supplemental Figure 1).
-Feuda *et al.* accept the results of some analyses and reject others based on posterior predictive (PP) analyses of model adequacy, which assess how well a model explains variation in the data (Bollback 2002).
+Feuda *et al.* accept the results of some analyses and reject others based on posterior predictive (PP) analyses of model adequacy, which assess how well a model explains variation in the data [@Bollback:2002to].
 They considered five different PP statistics that capture different types of variation in the data, such as the heterogeneity of states across taxa (PP taxon hetero in figures here) and sites (PP site hetero).
@@
 Of the six recoded analyses, five are unresolved.
-Only a single analysis for which PP scores are available provides strong support for Porifera-sister, the CAT-GTR+G analysis of the SR-6 [4] recoded Whelan [3] matrix.
+Only a single analysis for which PP scores are available provides strong support for Porifera-sister, the CAT-GTR+G analysis of the SR-6 [@Susko:2007ds] recoded Whelan [@Whelan:2015jj] matrix.
 Furthermore, this analysis does not have the best score according to any of the five PP statistics they considered (Supplemental Figure 2).
@@
 It should also be noted that none of the reanalyses of Chang matrix strongly contradict that study.
-Chang *et al.* [2] found strong support for Ctenophora-sister, and all of the reanalyses of this same matrix by Feuda *et al.* also have strong support for Ctenophora-sister or are unresolved.
+Chang *et al.* [@Chang:2015hl] found strong support for Ctenophora-sister, and all of the reanalyses of this same matrix by Feuda *et al.* also have strong support for Ctenophora-sister or are unresolved.
 
 Though their study does not provide strong support for Porifera-sister, the sensitivity of their results to recoding provides an opportunity to better understand and evaluate the impact of recoding more generally.
-This is important given the growing interest in recoding [1].
+This is important given the growing interest in recoding [@Feuda:2017ew].
 Feuda *et al.* hoped recoding would reduce potential artefacts due to differences across species in amino acid frequencies.
@@
 I created four new random recoding schemes by shuffling the amino acids in the SR-6 scheme (see Supplemental Methods and analysis code at https://github.com/caseywdunn/feuda_2017).
-When I applied each of these randomized codes to the Whelan matrix and analyzed them under the CAT-GTR+G model with phylobayes-mpi [6], I observed similar results as for the empirical SR-6 recoding.
+When I applied each of these randomized codes to the Whelan matrix and analyzed them under the CAT-GTR+G model with phylobayes-mpi [@Lartillot:2013fg], I observed similar results as for the empirical SR-6 recoding.
 Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Figure 1B).
@@
 This suggests that recoding may not be an effective tool for accommodating among-species differences in amino acid frequencies more generally.
-Compositional heterogeneity would be better addressed with models of molecular evolution that explicitly describe such differences [7,8], if progress can be made on the considerable computational challenges of such complex models.
+Compositional heterogeneity would be better addressed with models of molecular evolution that explicitly describe such differences [@Blanquart:2008gl, @Foster:2004tw], if progress can be made on the considerable computational challenges of such complex models.
 
@@
 CWD performed all analyses and wrote the manuscript.
-
-## References
-
-1.	Feuda, R., Dohrmann, M., Pett, W., Philippe, H., Rota-Stabelli, O., Lartillot, N., Wörheide, G., and Pisani, D.
-(2017).
-Improved Modeling of Compositional Heterogeneity Supports Sponges as Sister to All Other Animals.
-Curr.
-Biol.
-27, 3864–3870.e4.
-2.	Chang, E.S., Neuhof, M., Rubinstein, N.D., Diamant, A., Philippe, H., Huchon, D., and Cartwright, P.
-(2015).
-Genomic insights into the evolutionary origin of Myxozoa within Cnidaria.
-Proc.
-Natl.
-Acad.
-Sci.
-U.
-S.
-A.
-112, 14912–14917.
-3.	Whelan, N.V., Kocot, K.M., Moroz, L.L., and Halanych, K.M. (2015). Error, signal, and the placement of Ctenophora sister to all other animals. Proc. Natl. Acad. Sci. U. S. A. 112, 5773–5778.
-4.	Susko, E., and Roger, A.J. (2007). On reduced amino acid alphabets for phylogenetic inference. Mol. Biol. Evol. 24, 2139–2150.
-5.	Bollback, J.P. (2002). Bayesian model adequacy and choice in phylogenetics. Mol. Biol. Evol. 19, 1171–1180.
-6.	Lartillot, N., Rodrigue, N., Stubbs, D., and Richer, J. (2013). PhyloBayes MPI: phylogenetic reconstruction with infinite mixtures of profiles in a parallel environment. Syst. Biol. 62, 611–615.
-7.	Blanquart, S., and Lartillot, N. (2008). A site- and time-heterogeneous model of amino acid replacement. Mol. Biol. Evol. 25, 842–858.
-8.	Foster, P.G. (2004). Modeling compositional heterogeneity. Syst. Biol. 53, 485–495.
 
@@
 
-The original SR-6 recoding scheme is "`APST CW DEGN FHY ILMV KQR`" [4], where spaces separate amino acids that are placed into the same group.
+The original SR-6 recoding scheme is "`APST CW DEGN FHY ILMV KQR`" [@Susko:2007ds], where spaces separate amino acids that are placed into the same group.
 This recoding is one member of a family of recodings, each with a different number of groups, based on clustering of the JTT matrix.
-The other recoding used by Feuda *et al.*, KGB-6 and D-6, are based on different matrices and methods [1].
+The other recoding used by Feuda *et al.*, KGB-6 and D-6, are based on different matrices and methods [@Feuda:2017ew].
 
@@
 The only one of these eight analyses that provides strong support for Porifera-sister is not the most adequate analysis by any of the PP scores, and showed the poorest convergence according to Maxdiff.
+
+## References

### 2018-06-21 manuscript.rmd
word smithing

@@
 output:
-  github_document
+  github_document: default
+  pdf_document:
+    latex_engine: xelatex
+  word_document: default
 ---
@@
 
-Feuda *et al.* tackled a difficult phylogenetic problem -- placing the root of the animal phylogeny [@King:2017ie].
+Feuda *et al.* [@Feuda:2017ew] tackled a difficult phylogenetic problem -- placing the root of the animal phylogeny [@King:2017ie].
 In the past decade, some analyses have placed Ctenophora (comb jellies) and others Porifera (sponges) as the sister group to all other animals (Figure 1A).
@@
 Rather than model compositional heterogeneity, as their title suggests, this approach discards heterogeneous information so that much simpler models with fewer states can be applied.
-They report that posterior predictive (PP) analyses [@Bollback:2002to] indicate recoded analyses have better model adequacy than amino acid analyses, and "Porifera-sister was favored under all recoding strategies".
-Here I focus on two aspects of Feuda *et al.* First, I point out that many of their recoded analyses are actually unresolved, and that the analyses with the best PP scores do not provide strong support for Porifera-sister.
+They report that posterior predictive (PP) analyses [@Bollback:2002to] indicate 6-state recoded analyses have better model adequacy than 20-state amino acid analyses, and "Porifera-sister was favored under all recoding strategies".
+Here I focus on two aspects of Feuda *et al.* First, I point out that many of their recoded analyses are actually unresolved (*i.e.*, without strong support for either Porifera-sister or Ctenophora-sister), and that the analyses with the best posterior predictive scores do not provide strong support for Porifera-sister.
 Second, I present new analyses that show that the impact of recoding is largely due to discarding information, not accommodating variation in amino acid composition across lineages.
 These findings indicate that it is premature to accept Porifera-sister and reject Ctenophora-sister.
-The new analyses show that recoding can be a problematic method for addressing compositional variation in phylogenetic analyses.
+The new analyses also show that recoding can be a problematic method for addressing compositional variation in phylogenetic analyses.
 
@@
 Feuda *et al.* accept the results of some analyses and reject others based on posterior predictive (PP) analyses of model adequacy, which assess how well a model explains variation in the data [@Bollback:2002to].
-They considered five different PP statistics that capture different types of variation in the data, such as the heterogeneity of states across taxa (PP taxon hetero in figures here) and sites (PP site hetero).
+They considered five different posterior predictive statistics that capture different types of variation in the data, such as the heterogeneity of states across taxa ("PP taxon hetero" in figures here) and sites ("PP site hetero").
 From this they conclude that their "results strongly support sponges as the sister group of all other animals".
@@
 This conclusion does not follow from their own presented results.
-Only a single analysis with PP scores provides what could be considered strong support (&ge; `r posterior_prob_thresh` posterior probability) for Porifera-sister.
-Of the 32 analyses, PP scores were calculated for 16 (those for the full Whelan and Chan matrices).
-Based on PP scores, Feuda *et al.* reject eight of these that were conducted under the GTR+G model (which all have strong support for Ctenophora-sister).
+Only a single analysis with posterior predictive scores provides what could be considered strong support (> `r posterior_prob_thresh` posterior probability) for Porifera-sister.
+Of the 32 analyses, posterior predictive scores were calculated for 16 (those for the full Whelan and Chang matrices, but not for two Whelan matrices with reduced outgroup sampling).
+Based on posterior predictive scores, Feuda *et al.* reject eight of these that were conducted under the GTR+G model (which all have strong support for Ctenophora-sister).
 This leaves eight CAT-GTR+G analyses (Supplemental Figure 2).
@@
 Of the six recoded analyses, five are unresolved.
-Only a single analysis for which PP scores are available provides strong support for Porifera-sister, the CAT-GTR+G analysis of the SR-6 [@Susko:2007ds] recoded Whelan [@Whelan:2015jj] matrix.
-Furthermore, this analysis does not have the best score according to any of the five PP statistics they considered (Supplemental Figure 2).
+Only a single analysis for which posterior predictive scores are available provides strong support for Porifera-sister, the CAT-GTR+G analysis of the SR-6 [@Susko:2007ds] recoded Whelan [@Whelan:2015jj] matrix.
+Furthermore, this analysis does not have the best score according to any of the five posterior predictive statistics they considered (Supplemental Figure 2).
 The only statistic that stands out for this one analysis is that it has the highest maxdiff (Supplemental Figure 2), indicating that it did not converge as well as other analyses.
@@
 Similar results would suggest that the impact of recoding is due to discarding information.
-Here I focus on the single original analysis with a PP score that supports Porifera-sister, the CAT-GTR+G analysis of the SR-6 recoded Whelan data.
+Here I focus on their single analysis with a posterior predictive score that supports Porifera-sister, the CAT-GTR+G analysis of the SR-6 recoded Whelan data.
 I created four new random recoding schemes by shuffling the amino acids in the SR-6 scheme (see Supplemental Methods and analysis code at https://github.com/caseywdunn/feuda_2017).
@@
 
-These analyses indicate that the major impact of recoding on phylogenetic analyses is data reduction, not accomodation of compositional heterogeneity across species.
+These analyses indicate that the major impact of recoding on phylogenetic analyses is data reduction, not accommodation of compositional heterogeneity across species.
 This suggests that recoding may not be an effective tool for accommodating among-species differences in amino acid frequencies more generally.
@@
 (B) Each of the six plots presents one statistic, which include Posterior probability of Porifera-sister and the five Posterior Predictive (PP) statistics considered by Feuda *et al.* Within each plot, there are six lines for six different analyses.
-These six analyses are the published SR-6 analyses presented by Feuda *et al.* (SR6-published), reproduced analyses obtained by applying the same methods to the same data (SR6-reproduced), and four analyses based on randomized recoding matrices obtained by shuffling the SR-6 coding scheme (random-00 – random-03).
+These six analyses are the published SR-6 analyses presented by Feuda *et al.* (SR6-published), analyses obtained by applying the same methods to the same data to to confirm that I can reproduce their published results (SR6-reproduced), and four analyses based on randomized recoding matrices obtained by shuffling the SR-6 coding scheme (random-00 – random-03).
 Each analysis includes results for 20 states (the raw amino acid data, shown by the left point) and for 6 states (the 6-state recoded data, shown by the right point).
@@
 The 20 state matrices are the same across all analyses since they are not recoded.
-Since all the 20 state matrices are all the same, variation between 20-state results (as in the left side of each pane of Figure 1B) give insight into the technical variance of the inference process.
+Since all the 20 state matrices are the same, variation between 20-state results (as in the left side of each pane of Figure 1B) give insight into the technical variance of the inference process.
 
@@
 Analyses were run for 1000 generations, and a `r burnin` generation burnin applied.
-The resulting tree files and PP scores were parsed for display with the code in https://github.com/caseywdunn/feuda_2017/manuscript.rmd .
+The resulting tree files and posterior predictive scores were parsed for display with the code in https://github.com/caseywdunn/feuda_2017/manuscript.rmd .
 
 The statistics presented in Supplemental Figures 1 and 2 were parsed from the Feuda *et al.* manuscript into the file `tidy_table_3.tsv` and rendered for display with the code on the source file for this document.
+
+\pagebreak
 
@@
 ```
-**Supplemental Figure 1.**  A grapchical representation of the posterior probabilities for the 32 analyses presented by Feuda *et al.* in their Table 3.
-Cells are color coded by whether posterior probability is &ge; `r posterior_prob_thresh` for Porifera-sister, &ge; `r posterior_prob_thresh` for Ctenophora-sister, or neither (unresolved).
-Posterior predictive (PP) statistics were estimated for the 16 analyses in the top two rows of this figure.
+**Supplemental Figure 1.**  A graphical representation of the posterior probabilities for the 32 analyses presented by Feuda *et al.* in their Table 3.
+Cells are color coded by whether posterior probability is > `r posterior_prob_thresh` for Porifera-sister, > `r posterior_prob_thresh` for Ctenophora-sister, or neither (unresolved).
+Posterior predictive (PP) statistics were estimated for the 16 analyses in the top two rows of this figure (the Chang and full Whelan matrices), but not the bottom two (the Whelan matrices with reduced outgroup sampling).
+
+\pagebreak
 
@@
 Cells are color coded as in Supplemental Figure 1.
-Only one of these analyses, the SR6 coding of the Whelan matrix, has &ge; `r posterior_prob_thresh` support for Porifera-sister.
+Only one of these analyses, the SR6 coding of the Whelan matrix, has > `r posterior_prob_thresh` support for Porifera-sister.
 The 20-state and 6-state points on the plots in Figure 1 of the main text correspond to the 20 and SR6 Whelan cells shown here.
-The presented statistics for these cells are posterior probability of Porifera-sister, Maxdiff (with lower scores indicating better convergence of runs), and five PP statistics (where lower absolute value indicates better model adequacy).
-The only one of these eight analyses that provides strong support for Porifera-sister is not the most adequate analysis by any of the PP scores, and showed the poorest convergence according to Maxdiff.
+The presented statistics for these cells are posterior probability of Porifera-sister, Maxdiff (with lower scores indicating better convergence of runs), and five posterior predictive statistics (where lower absolute value indicates better model adequacy).
+The only one of these eight analyses that provides strong support for Porifera-sister is not the most adequate analysis by any of the posterior predictive scores, and showed the poorest convergence according to Maxdiff.
+
+\pagebreak
 

### 2018-06-21 manuscript.rmd
removed link

@@
 output:
-  github_document: default
   pdf_document:
     latex_engine: xelatex
+  github_document: default
   word_document: default
@@
 Analyses were run for 1000 generations, and a `r burnin` generation burnin applied.
-The resulting tree files and posterior predictive scores were parsed for display with the code in https://github.com/caseywdunn/feuda_2017/manuscript.rmd .
+The resulting tree files and posterior predictive scores were parsed for display with the code in `manuscript.rmd`.
 
-The statistics presented in Supplemental Figures 1 and 2 were parsed from the Feuda *et al.* manuscript into the file `tidy_table_3.tsv` and rendered for display with the code on the source file for this document.
+The statistics presented in Supplemental Figures 1 and 2 were parsed from the Feuda *et al.* manuscript into the file `tidy_table_3.tsv` and rendered for display with the code in `manuscript.rmd`.
 

### 2018-09-18 manuscript.rmd
shortened text

@@
 output:
+  github_document: default
   pdf_document:
     latex_engine: xelatex
-  github_document: default
   word_document: default
@@
 In the past decade, some analyses have placed Ctenophora (comb jellies) and others Porifera (sponges) as the sister group to all other animals (Figure 1A).
-Feuda *et al.* [@Feuda:2017ew] present new phylogenetic analyses that they claim provide conclusive support for the Porifera-sister hypothesis.
+Feuda *et al.* [@Feuda:2017ew] present new phylogenetic analyses that they claim provide strong support for Porifera-sister.
 Their analyses consider datasets from studies by Chang *et al.* [@Chang:2015hl] and Whelan *et al.* [@Whelan:2015jj], both of which found support for Ctenophora-sister.
@@
 Here I focus on two aspects of Feuda *et al.* First, I point out that many of their recoded analyses are actually unresolved (*i.e.*, without strong support for either Porifera-sister or Ctenophora-sister), and that the analyses with the best posterior predictive scores do not provide strong support for Porifera-sister.
-Second, I present new analyses that show that the impact of recoding is largely due to discarding information, not accommodating variation in amino acid composition across lineages.
+Second, I present new analyses that show the impact of recoding is largely due to discarding information, not accommodating variation in amino acid composition.
 These findings indicate that it is premature to accept Porifera-sister and reject Ctenophora-sister.
-The new analyses also show that recoding can be a problematic method for addressing compositional variation in phylogenetic analyses.
+They also show that recoding can be a problematic method for addressing compositional variation.
 
@@
 Feuda *et al.* accept the results of some analyses and reject others based on posterior predictive (PP) analyses of model adequacy, which assess how well a model explains variation in the data [@Bollback:2002to].
-They considered five different posterior predictive statistics that capture different types of variation in the data, such as the heterogeneity of states across taxa ("PP taxon hetero" in figures here) and sites ("PP site hetero").
+They considered five different posterior predictive statistics that capture different types of variation in the data.
 From this they conclude that their "results strongly support sponges as the sister group of all other animals".
@@
 Only a single analysis with posterior predictive scores provides what could be considered strong support (> `r posterior_prob_thresh` posterior probability) for Porifera-sister.
-Of the 32 analyses, posterior predictive scores were calculated for 16 (those for the full Whelan and Chang matrices, but not for two Whelan matrices with reduced outgroup sampling).
+Of the 32 analyses, posterior predictive scores were calculated for 16 (those for the full Whelan and Chang matrices).
 Based on posterior predictive scores, Feuda *et al.* reject eight of these that were conducted under the GTR+G model (which all have strong support for Ctenophora-sister).
@@
 The only statistic that stands out for this one analysis is that it has the highest maxdiff (Supplemental Figure 2), indicating that it did not converge as well as other analyses.
-It should also be noted that none of the reanalyses of Chang matrix strongly contradict that study.
-Chang *et al.* [@Chang:2015hl] found strong support for Ctenophora-sister, and all of the reanalyses of this same matrix by Feuda *et al.* also have strong support for Ctenophora-sister or are unresolved.
 
@@
 They interpreted the fact that their analyses are sensitive to recoding as evidence that such an artefact exists and that they successfully addressed it by recoding.
-An alternative hypothesis is that recoding impacts phylogenetic analyses simply because it discards so much information.
+An alternative hypothesis is that recoding impacts phylogenetic analyses because it discards so much information.
 These two hypotheses can be tested by applying new recoding schemes that also reduce twenty states down to six, but are based on random grouping rather than empirical frequencies of amino acid exchange.
@@
 
-These analyses indicate that the major impact of recoding on phylogenetic analyses is data reduction, not accommodation of compositional heterogeneity across species.
-This suggests that recoding may not be an effective tool for accommodating among-species differences in amino acid frequencies more generally.
+These analyses suggest that the major impact of recoding on phylogenetic analyses is data reduction, not accommodation of compositional heterogeneity across species.
+This indicates that recoding may not be an effective tool for accommodating among-species differences in amino acid frequencies.
 Compositional heterogeneity would be better addressed with models of molecular evolution that explicitly describe such differences [@Blanquart:2008gl, @Foster:2004tw], if progress can be made on the considerable computational challenges of such complex models.
